### 2017-03-06 201703-aha-poster-2/aha-2017-cfa-slides.Rnw
initial commit programs

% ser-2017-aha-slcs.Rnw

%---------------------------------------------------------------
% Preamble
% --------------------------------------------------------------
%
% NOTE: See rice-sample.tex written by Daina Chiba at Rice University for formatting and preamble code that I copied, http://ricebeamer.dynaman.net/
\documentclass[pdf]{beamer}

\usetheme{Frankfurt}
\usecolortheme{beetle}
\setbeamersize{text margin left=10pt,text margin right=10pt} %set margin sizes

\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{bm}
\usepackage{blindtext}
\usepackage{scrextend}
\addtokomafont{labelinglabel}{\sffamily}
\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{caption}

\usepackage{graphicx}% http://ctan.org/pkg/graphicx

\setbeamercolor{bibliography entry title}{fg=black,bg=black}% see http://tex.stackexchange.com/questions/71352/beamer-undefined-color-local-structure
\setbeamercolor{bibliography entry location}{fg=white}
\setbeamercolor{bibliography entry note}{fg=white}
\setbeamertemplate{caption}[numbered]

% got from http://tex.stackexchange.com/questions/48023/mimic-bibtex-apalike-with-biblatex-biblatex-apa-broken
\PassOptionsToPackage{
        style=numeric,
        hyperref=true,
        backend=bibtex,
        maxbibnames=1,
        firstinits=true,
        uniquename=init,
        maxcitenames=2,
        parentracker=true,
        url=false,
        doi=true,
        isbn=false,
        eprint=false,
        backref=false,
            }{biblatex}
% see the following link for info on biblatex sort order issue: 
% http://tex.stackexchange.com/questions/51434/biblatex-citation-order
\usepackage[natbib=true, sorting=none, style=numeric, backend=biber]{biblatex}
\addbibresource{bib1}
\renewcommand*{\bibfont}{\scriptsize}

% see http://tex.stackexchange.com/questions/43083/author-year-abbr-journal-name-as-citation-style
\DeclareCiteCommand{\longcite}{(}{%
    \printnames[author]{author}, \printfield{year})}{}{}%

\newcommand{\customcite}[1]{\citeauthor{#1}, \citetitle{#1}, \citeyear{#1}}
\newcommand{\customcitetwo}[1]{\citeauthor{#1}, \citeyear{#1}}

\renewcommand{\footnotesize}{\tiny}  % change font size of citation
\renewcommand\multicitedelim{\addsemicolon\space} % This doesn't seem to work, but see http://tex.stackexchange.com/questions/167665/multiple-references-with-footfullcite

%\usepackage{fontspec} % have to compile with XeLaTeX
%\setmainfont{Arial}
\usepackage[T1]{fontenc}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault} % get something like Arial

\usepackage{amsmath,amsthm, amssymb, latexsym}
\usepackage{wrapfig}

\usepackage{array,booktabs,tabularx}
\newcolumntype{Z}{>{\centering\arraybackslash}X} % centered tabularx columns

\usepackage[shortlabels]{enumitem}
% \setlist[description]{style=nextline}
\setlist[itemize]{leftmargin=0.3in, labelindent=16pt, label=$\Diamond$}

% set bullet points in itemize
% see http://tex.stackexchange.com/questions/184080/using-beamer-bullets-in-poster
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,decorations.markings,snakes,shapes}
\tikzset{
font={\fontsize{14pt}{16}\selectfont}}

\usepackage[dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{colortbl}
\usepackage{adjustbox}

\usepackage[framemethod=TikZ]{mdframed}
\mdfdefinestyle{MyFrame}{%
    linecolor=carolinablue,
    outerlinewidth=4pt,
    roundcorner=20pt,
    innerrightmargin=20pt,
    innerleftmargin=20pt,
    backgroundcolor=gray!10!white}

% set three column widths in poster
\newlength{\columnheight}
\setlength{\columnheight}{105cm}
\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\sepwid}{0.024\paperwidth}
\setlength{\onecolwid}{0.31\paperwidth}
\setlength{\twocolwid}{0.31\paperwidth}
\setlength{\threecolwid}{0.31\paperwidth}

<<mainSettings, echo=FALSE, include=FALSE>>=
opts_chunk$set(echo=F, comment=NA, message=T, results = 'hide', cache=F)

library(knitr)
library(data.table)
library(readxl)
library(ggplot2)
library(plyr)
library("MplusAutomation")
library(ztable)
options(ztable.type="latex")
library(Hmisc)
library(htmlTable)
@

% --------------------------------------------------------------------------------------% 
% Title, author, date, etc.
% --------------------------------------------------------------------------------------% 
% see http://tex.stackexchange.com/questions/9740/how-can-i-add-vertical-space-to-a-beamercolorbox-to-make-it-align-with-another-o
\title{Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups: United States, 2003-2012} 
\author[vonholle@unc.edu]{Ann Von Holle, Kari E. North, UNC, Chapel Hill, NC}
\institute{UNC, Chapel Hill}
\titlegraphic{\includegraphics[scale=0.25]{unc-logo.png}} %this is the path to your logo

% -------------------------------------------------------------------------------------%
% Contents
% -------------------------------------------------------------------------------------%

\begin{document}

\begin{frame}
\titlepage
\centering
\end{frame}

% \itemsep0.3em
\section{Introduction}
\subsection{Introduction}

\begin{frame}{Introduction}
          \begin{itemize}
          \normalsize  
          \item  High density lipoprotein cholesterol (HDL-C), low density lipoprotein cholesterol (LDL-C) and triglycerides (TG) are common measures of dyslipidemia.
              \begin{itemize}
                \normalsize \item Common analytic practice includes separate evaluations of each lipoprotein when evaluating dyslipidemia. 
                \item An alternative is to specify one latent factor, which predicts lipid values and incorporates their covariance structure.
                \end{itemize}
            \item Before using the latent factor, important to determine if underlying structure of dyslipidemia construct is similar across groups (measurement invariance). 
            \begin{itemize}
              \normalsize \item For example, do racial/ethnic groups have the same construct?
              \end{itemize}
            \end{itemize}
\end{frame}

        \section{Aims}
        \subsection{Aims}

        \begin{frame}{Aims}

          \begin{description}
          \item[Aim 1] Estimate dyslipidemia in a U.S. representative sample (NHANES) as one latent factor with three indicators (HDL-C, LDL-C, TG) via confirmatory factor analysis (CFA).
            \item[Aim 2] Test similarity of latent factor parameters (measurement invariance)  \footfullcite{van_de_schoot_checklist_2012} across age and racial/ethnic groups.
            \end{description}

        \end{frame}

        % -----------------------------------------------------------
        % 1-3
        % -----------------------------------------------------------
        \section{Methods}
        \subsection{Methods}

        % Figure showing{ CFA
        % %%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Confirmatory Factor Analysis (CFA) Model}

          \tikzset{
            gla/.style={
          , line width=0.5mm, arrows={-Stealth[angle=45:12pt, scale=1, black, fill=black]}}}

          \tikzset{
            dgla/.style={
          , line width=0.5mm, shorten >=1pt, arrows={Stealth-Stealth[angle=45:10pt, scale=1, black, fill=black]}}}

          \tikzset{
    ultra thin/.style= {line width=0.1pt},
    very thin/.style=  {line width=0.2pt},
    thin/.style=       {line width=0.4pt},% thin is the default
    rec/.style=        {shape=rectangle, line width=0.5mm, draw},% thin is the default
    semithick/.style=  {line width=0.6pt},
    thick/.style=      {line width=0.8pt},
    very thick/.style= {line width=1.2pt},
    ultra thick/.style={line width=1.6pt}
}

\centering
        \begin{figure}
          	\begin{tikzpicture}
          		\node[line width=0.5mm, draw, ellipse, minimum height=1cm] (dys) {\large Dyslipidemia};

          		\node[above left=of dys, rec] (hdl) {\normalsize HDL-C};
          		\node[above=of dys, rec] (ldl) {\normalsize LDL-C};
          		\node[above right=of dys, rec] (tg) {\normalsize TG};

          		\node[above=of hdl, draw=none, fill=none] (e1) {\normalsize $\epsilon_1$};
          		\node[above=of ldl, draw=none, fill=none] (e2) {\normalsize $\epsilon_2$};
          		\node[above=of tg, draw=none, fill=none] (e3) {\normalsize $\epsilon_3$};

          		\draw [->, gla] (e1) to (hdl);
          		\draw [->, gla] (e2) to (ldl);
          		\draw [->, gla] (e3) to (tg);

          		\draw [->, gla] (dys) to node [anchor=east] {\large$\lambda_1$} (hdl)  ;
          		\draw [->, gla] (dys) to node [anchor=west] {\large$\lambda_2$} (ldl);
          		\draw [->, gla] (dys) to node [anchor=west] {\large$\lambda_3$} (tg);
          		\draw [<->, dgla, bend left=60]  (e1) to node [anchor=north] {\normalsize$\epsilon_{12}$} (e2);
          	\end{tikzpicture}
      \end{figure}
\end{frame}

\begin{frame}{Equations for CFA\footfullcite{brown_confirmatory_2015}}

\centering
% Equations here
\begin{array}{rcl} 
  \textrm{HDL-C} & = & \mu_1 + \lambda_1 \eta_1 + \epsilon_1 \\ 
  \textrm{LDL-C} & = & \mu_2 + \lambda_2 \eta_1 + \epsilon_2 \\ 
  \textrm{TG-C} & = & \mu_3 + \lambda_3 \eta_1 + \epsilon_3 \\ 
\end{array}

\vskip1em

Parameters:
\begin{mdframed}
\begin{description}
  \item[Factor loadings] $\lambda$, represent the association between lipids and the latent factor for dyslipidemia, $\eta_1$.
  \item[Intercepts] $\mu$, represent the means of the indicators, lipids.
  \end{description}
\end{mdframed}

\end{frame}

\begin{frame}{Four models$^a$}

\begin{itemize}
%see http://latex.wikia.com/wiki/List_of_LaTeX_symbols
  \item[\Rightarrow]\normalsize Evaluate model fit across year and racial/ethnic groups in four models, holding different parameters constant.
  \end{itemize}

\centering

  \begin{description}[align=left]
  \normalsize
    \item [Model 1]\raggedright Test metric invariance: equal factor loadings but different intercepts across groups.
    \item [Model 2]\raggedright Test intercept-only invariance: different factor loadings but equal intercepts across groups
    \item [Model 3]\raggedright Test scalar invariance: equal factor loadings and equal intercepts across groups
    \item [Model 4]\raggedright Full uniqueness invariance: equal loadings, intercept and variance across all groups
  \end{description}

\indent \raggedright \scriptsize $^a$  All models include body mass index (BMI) and age covarying with factor (not shown in figure).
      \end{frame}

        % -----------------------------------------------------------
        % 2-2
        % -----------------------------------------------------------
      \section{Results, 12-19 years}
      \subsection{Results, 12-19 years}

        % Read in code from mplus results web page development

        <<code1, cache=F>>=
          read_chunk("../mplus_results.R")
        @

        <<read1, eval=F>>=
          #NOTE: only run this as needed (with new data). Takes long time to run.
        @

        <<data-handle1>>=
        @

        <<data-handle2>>=
        @

        <<fit1>>=
        @

% see http://tex.stackexchange.com/questions/94078/resize-table-and-caption
\begin{frame}{Model fit information by groups}
\centering

\begin{minipage}[h]{1\textwidth}  
        \begin{adjustbox}{width=\linewidth} 
<<printz-1, results='asis', eval=T>>=
        print(z,type="latex")
        @
        \end{adjustbox}
\end{minipage}

\end{frame}

\begin{frame}{Best fit selections for racial/ethnic groups}

\begin{itemize} 
  \item Model 1, allowing intercepts to vary, offers best fit.
    \begin{itemize} \normalsize
      \item Acceptable levels include Root Mean Square Error of Approximation (RMSEA) < 0.05, Comparative Fit Index (CFI) = 0.95 and lowest Bayesian Information Criterion (BIC) of all models.
      \end{itemize}
    \item Evidence for metric noninvariance  ($\chi^2$(df)=155 (8), p-value<0.001) $\Rightarrow$ Cannot make meaningful comparisons across racially and ethnically diverse populations.

  \end{itemize}
\end{frame}

\begin{frame}{Best fit selections for time groups}

  \begin{itemize}
    \item Model 1, allowing intercepts to vary, offers best fit.
    \begin{itemize} \normalsize
      \item RMSEA < 0.05 and lowest Bayesian Information Criterion (BIC) of all models.
      \end{itemize}
    \item Evidence for metric noninvariance ($\chi^2$(df)=155 (12), p-value<0.001)  $\Rightarrow$ Cannot make meaningful comparisons across time.
    \end{itemize}
\end{frame}

% Second table here
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<loadings1>>=
@

<<combo1>>=
@

<<print-combo-race,  warning=TRUE>>=
@

\begin{frame}{Model 1 standardized parameters by race/ethnic groups}

\centering
\begin{minipage}[h]{0.9\textwidth}                          % minipage spans half the textwidth
        \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
<<printz-3, results='asis', eval=T>>=
        print(z,type="latex")
        @
        \end{adjustbox}
\end{minipage}
\end{frame}

\begin{frame}{Latent factor characteristics by racial/ethnic groups}

          \begin{itemize}
            \item \underline{Factor loadings}: A one unit factor increase represents less favorable lipid value $\Rightarrow$ an increase in LDL-C (0.4) and TG (0.6) accompanied by a decrease in HDL ($\sim$ -0.6).
            \item \underline{Intercepts}: Non-Hispanic Black group only one with positive HDL-C (0.132) and lowest TG (-1.078) relative to center of multivariate distribution.
          \end{itemize}
\end{frame}

<<print-combo-year,  warning=TRUE>>=
@

\begin{frame}{Model 1 standardized parameters by year groups}

\centering
\begin{minipage}[h]{0.8\textwidth}                          % minipage spans half the textwidth
        \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
<<printz-4, results='asis', eval=T>>=
        print(z,type="latex")
        @
        \end{adjustbox}
\end{minipage}
\end{frame}

% NOTE: not enough space here for gender. Remove to allow more
% discussion of results in poster. keep for slides

\begin{frame}{Fit and parameters for groups by sex}

\vskip-1em

% Run code for gender groups (from mplus_results.R)
<<fit-gender>>=
@
\centering
\begin{adjustbox}{width=0.7\linewidth} 
%\caption*{Model fit}
<<printf-gender, results='asis', eval=T>>=
        print(z,type="latex",
              caption="Model fit")
@
\end{adjustbox}

\vskip1em

<<print-combo-gender>>=
@
\centering
\begin{adjustbox}{width=0.7\linewidth}
%\caption*{Model 1 parameters}
<<printp-gender, results='asis', eval=T>>=
        print(z,type="latex",
              caption="Model 1 parameters")
@
\end{adjustbox}

\vskip1em

\raggedright \footnotesize Invariance testing indicates evidence for metric non-invariance when testing difference between a metric model (model 1) and a full invariance model (model 4) in which there are full equality constraints across groups ($\chi^2$ = 68.5 (df=2); p-value<0.001).

% NOTE: below is column approach, but too small to read

% \begin{columns}[T] % align columns
% \begin{column}{.48\textwidth}
% 
% %Left Part -------------------
% 
% % Run code for gender groups (from mplus_results.R)
% <<fit-gender>>=
% @
% \centering
% \begin{minipage}[h]{1\textwidth}  
% \begin{adjustbox}{width=\linewidth} 
% %\caption*{Model fit}
% <<printf-gender, results='asis', eval=T>>=
%         print(z,type="latex",
%               caption="Model fit")
% @
% \end{adjustbox}
% \end{minipage}
% 
% 
% \end{column}%
% 
% \hfill%
% \begin{column}{.48\textwidth}
% 
% 
% % Right Part --------------------
% 
% <<print-combo-gender>>=
% @
% \centering
% \begin{minipage}[h]{1\textwidth}  
% \begin{adjustbox}{width=\linewidth}
% %\caption*{Model 1 parameters}
% <<printp-gender, results='asis', eval=T>>=
%         print(z,type="latex",
%               caption="Model 1 parameters")
% @
% \end{adjustbox}
% \end{minipage}
% 
% \end{column}%
% \end{columns}

\end{frame}

% >>>>>>>>>>>>>>>>>>>>>>>>>>>

\begin{frame}{Model fitting process}
          \begin{itemize}
          \item Covariance between HDL-C and TG no longer significant when accounting for BMI covarying with latent lipid factor.
          \item Age positive covariance with lipid factor.
          \item BMI positive covariance with lipid factor.
          \end{itemize}
\end{frame}

\section{Summary}
\subsection{Summary}

\begin{frame}{Summary} 
\Large
            \begin{itemize}
                \item 3-indicator latent factor can function as a parsimonious and well-fitting measure for dyslipidemia in structural equation models.
                \item []
                \item Important not to compare latent factor characteristics across groups by race/ethnic, time or sex status given metric measurement non-invariance.
            \end{itemize}
\end{frame}

\begin{frame}

Contact information: vonholle at unc dot edu

Further project information: https://avonholle.github.io/cfa-lipids/

\end{frame}

\end{document}

### 2017-03-06 201703-aha-poster-2/aha-2017-cfa.Rnw
initial commit programs

--- 2017-03-06 201703-aha-poster-2/aha-2017-cfa-slides.Rnw
@@
-% ser-2017-aha-slcs.Rnw
+% aha-2017-slcs.Rnw
 
@@
 % NOTE: See rice-sample.tex written by Daina Chiba at Rice University for formatting and preamble code that I copied, http://ricebeamer.dynaman.net/
-\documentclass[pdf]{beamer}
+\documentclass[final]{beamer}
+\usepackage[orientation=landscape, size=custom, width=121.92, height=106.68, scale=1.7]{beamerposter}  % this matches 4 feet (48 inches) by 3.5 feet (42 inches).
+Will get from phdposters.com 1 inch per 2.54 cm
 
-\usetheme{Frankfurt}
-\usecolortheme{beetle}
-\setbeamersize{text margin left=10pt,text margin right=10pt} %set margin sizes
-
+\mode<presentation>{\usetheme{UNC5}}
 \usepackage[english]{babel}
@@
 \usepackage{csquotes}
-\usepackage{booktabs}
-\usepackage{caption}
-
-\usepackage{graphicx}% http://ctan.org/pkg/graphicx
 
 \setbeamercolor{bibliography entry title}{fg=black,bg=black}% see http://tex.stackexchange.com/questions/71352/beamer-undefined-color-local-structure
-\setbeamercolor{bibliography entry location}{fg=white}
-\setbeamercolor{bibliography entry note}{fg=white}
 \setbeamertemplate{caption}[numbered]
@@
 
-% see http://tex.stackexchange.com/questions/43083/author-year-abbr-journal-name-as-citation-style
-\DeclareCiteCommand{\longcite}{(}{%
-    \printnames[author]{author}, \printfield{year})}{}{}%
-
-\newcommand{\customcite}[1]{\citeauthor{#1}, \citetitle{#1}, \citeyear{#1}}
-\newcommand{\customcitetwo}[1]{\citeauthor{#1}, \citeyear{#1}}
-
-\renewcommand{\footnotesize}{\tiny}  % change font size of citation
-\renewcommand\multicitedelim{\addsemicolon\space} % This doesn't seem to work, but see http://tex.stackexchange.com/questions/167665/multiple-references-with-footfullcite
-
 %\usepackage{fontspec} % have to compile with XeLaTeX
@@
 % \setlist[description]{style=nextline}
-\setlist[itemize]{leftmargin=0.3in, labelindent=16pt, label=$\Diamond$}
+\setlist[itemize]{leftmargin=0.6in, labelindent=16pt, label=$\Diamond$}
 
@@
 \author[vonholle@unc.edu]{Ann Von Holle, Kari E. North, UNC, Chapel Hill, NC}
-\institute{UNC, Chapel Hill}
-\titlegraphic{\includegraphics[scale=0.25]{unc-logo.png}} %this is the path to your logo
+\institute{UNC}
+\titlegraphic{unc-black.eps} %this is the path to your logo
 
@@
 
-\begin{frame}
-\titlepage
-\centering
-\end{frame}
+\begin{frame}[t]
 
+  \begin{columns}[T] % t instead of T or c means columns start at top
+
+    % ---------------------------------------------------------%
+    % Set up 1st column
+    % ---------------------------------------------------------%
+    \begin{column}{\onecolwid}
+    \begin{beamercolorbox}[wd=\textwidth]{postercolumn}
+    % fill each column with content
+        % -----------------------------------------------------------
+        % 1-1 (first column's first block
+        % -----------------------------------------------------------
+        % fill each column with content
+        \begin{block}{Introduction}
+
+          \begin{itemize}
 %          \itemsep0.3em
-\section{Introduction}
-\subsection{Introduction}
-
-\begin{frame}{Introduction}
-          \begin{itemize}
           \normalsize  
@@
             \end{itemize}
-\end{frame}
 
-        \section{Aims}
-        \subsection{Aims}
+        \end{block}
+        \vskip1ex
 
-        \begin{frame}{Aims}
+        % -----------------------------------------------------------
+        % 1-2
+        % -----------------------------------------------------------
+        \begin{block}{Aims}
 
-          \begin{description}
-          \item[Aim 1] Estimate dyslipidemia in a U.S. representative sample (NHANES) as one latent factor with three indicators (HDL-C, LDL-C, TG) via confirmatory factor analysis (CFA).
-            \item[Aim 2] Test similarity of latent factor parameters (measurement invariance)  \footfullcite{van_de_schoot_checklist_2012} across age and racial/ethnic groups.
-            \end{description}
+          \begin{enumerate}[1.]
+          \item Estimate dyslipidemia in a U.S. representative sample (NHANES) as one latent factor with three indicators (HDL-C, LDL-C, TG) via confirmatory factor analysis (CFA).
+            % \begin{itemize}
+            %   \item Compare model fit using common indicators.
+            %   \end{itemize}
+            \item Test similarity of latent factor parameters (measurement invariance) across age and racial/ethnic groups.
+            % \begin{itemize}
+            %   \normalfont \item Groups include racial/ethnic status and time (years).
+            %   \end{itemize}
+            \end{enumerate}
 
-        \end{frame}
+        \end{block}
+        \vskip1ex
 
@@
         % -----------------------------------------------------------
-        \section{Methods}
-        \subsection{Methods}
+        \begin{block}{Methods: CFA}
 
-        % Figure showing{ CFA
+        % Figure showing CFA
         % %%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}{Confirmatory Factor Analysis (CFA) Model}
 
@@
             gla/.style={
-          , line width=0.5mm, arrows={-Stealth[angle=45:12pt, scale=1, black, fill=black]}}}
+          , line width=1mm, arrows={-Stealth[angle=45:12pt, scale=2, black, fill=black]}}}
 
@@
             dgla/.style={
-          , line width=0.5mm, shorten >=1pt, arrows={Stealth-Stealth[angle=45:10pt, scale=1, black, fill=black]}}}
+          , line width=1mm, shorten >=1pt, arrows={Stealth-Stealth[angle=45:10pt, scale=1.5, black, fill=black]}}}
 
@@
     thin/.style=       {line width=0.4pt},% thin is the default
-    rec/.style=        {shape=rectangle, line width=0.5mm, draw},% thin is the default
+    rec/.style=        {shape=rectangle, line width=1mm, draw},% thin is the default
     semithick/.style=  {line width=0.6pt},
@@
 
+\vskip-1ex
 \centering
         \begin{figure}
+           \scalebox{1.2}{
           	\begin{tikzpicture}
-          		\node[line width=0.5mm, draw, ellipse, minimum height=1cm] (dys) {\large Dyslipidemia};
 
-          		\node[above left=of dys, rec] (hdl) {\normalsize HDL-C};
-          		\node[above=of dys, rec] (ldl) {\normalsize LDL-C};
-          		\node[above right=of dys, rec] (tg) {\normalsize TG};
+          		\node[line width=1mm, draw, ellipse, minimum height=1cm] (dys) {\large Dyslipidemia};
+
+          		\node[above left=of dys, xshift=2cm, yshift=4cm, rec] (hdl) {\normalsize HDL-C};
+          		\node[above=of dys, yshift=4cm, rec] (ldl) {\normalsize LDL-C};
+          		\node[above right=of dys, xshift=-2cm, yshift=4cm, rec] (tg) {\normalsize TG};
 
@@
           		\draw [<->, dgla, bend left=60]  (e1) to node [anchor=north] {\normalsize$\epsilon_{12}$} (e2);
+
           	\end{tikzpicture}
+           }
       \end{figure}
-\end{frame}
 
-\begin{frame}{Equations for CFA\footfullcite{brown_confirmatory_2015}}
-
-\centering
+Equations for CFA:\left\{
 % Equations here
@@
 
-\vskip1em
-
-Parameters:
-\begin{mdframed}
-\begin{description}
-  \item[Factor loadings] $\lambda$, represent the association between lipids and the latent factor for dyslipidemia, $\eta_1$.
-  \item[Intercepts] $\mu$, represent the means of the indicators, lipids.
-  \end{description}
-\end{mdframed}
-
-\end{frame}
-
-\begin{frame}{Four models$^a$}
-
 \begin{itemize}
@@
 
+        \end{block}
+
+  \end{beamercolorbox}
+  \end{column}
+    % ---------------------------------------------------------%
+    % end the 1st column
+    % ---------------------------------------------------------%
+
+% ---------------------------------------------------------%
+% Set up 2nd column
+% ---------------------------------------------------------%
+
+\begin{column}{\twocolwid}
+\begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
+
+        % -----------------------------------------------------------
+        % 2-1
+        % -----------------------------------------------------------
+      \begin{block}{Methods, cont...}
+
+\begin{itemize}
+  \item Factor loadings, $\lambda$, represent the association between lipids and the latent factor for dyslipidemia, $\eta_1$.
+  \item Intercepts, $\mu$, represent the means of the indicators, lipids.
+  \end{itemize}
+
 \centering
-
+\textbf{Four models$^a$}
+\begin{mdframed}[style=MyFrame]
   \begin{description}[align=left]
@@
     \end{description}
-
+\end{mdframed}
 \indent \raggedright \scriptsize $^a$  All models include body mass index (BMI) and age covarying with factor (not shown in figure).
-      \end{frame}
+      \end{block}
+      \vskip1ex
 
@@
         % -----------------------------------------------------------
-      \section{Results, 12-19 years}
-      \subsection{Results, 12-19 years}
+      \begin{block}{Results, 12-19 years}
 
@@
 % see http://tex.stackexchange.com/questions/94078/resize-table-and-caption
-\begin{frame}{Model fit information by groups}
+
 \centering
+Model fit information
 
-\begin{minipage}[h]{1\textwidth}  
-        \begin{adjustbox}{width=\linewidth} 
+\begin{mdframed}[style=MyFrame]
+\begin{minipage}[h]{1\textwidth}  % minipage spans half the textwidth
+        \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
 <<printz-1, results='asis', eval=T>>=
@@
 \end{minipage}
+\end{mdframed}
 
-\end{frame}
-
-\begin{frame}{Best fit selections for racial/ethnic groups}
-
+\textbf{Racial/ethnic groups}
 \begin{itemize} 
@@
       \item Acceptable levels include Root Mean Square Error of Approximation (RMSEA) < 0.05, Comparative Fit Index (CFI) = 0.95 and lowest Bayesian Information Criterion (BIC) of all models.
+%  \item No evidence model 1 different than configural model relaxing all parameter assumptions ($\chi^2$=9.12, p-value=0.33, df=8).
       \end{itemize}
-    \item Evidence for metric noninvariance  ($\chi^2$(df)=155 (8), p-value<0.001) $\Rightarrow$ Cannot make meaningful comparisons across racially and ethnically diverse populations.
+  \end{itemize}
 
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Best fit selections for time groups}
-
+\textbf{Time groups}
   \begin{itemize}
@@
     \begin{itemize} \normalsize
-      \item RMSEA < 0.05 and lowest Bayesian Information Criterion (BIC) of all models.
+      \item RMSEA < 0.05, CFI = 0.93 and lowest Bayesian Information Criterion (BIC) of all models.
+      %\item Evidence suggests metric and scalar invariance.% (all $\chi^2$ tests against configural model not significant).
       \end{itemize}
-    \item Evidence for metric noninvariance ($\chi^2$(df)=155 (12), p-value<0.001)  $\Rightarrow$ Cannot make meaningful comparisons across time.
     \end{itemize}
-\end{frame}
 
@@
 
+\end{block}
+
+\end{beamercolorbox}
+\end{column}
+
+% ---------------------------------------------------------%
+% end the 2nd column
+
+% ---------------------------------------------------------%
+% Set up 3rd column
+% ---------------------------------------------------------%
+
+\begin{column}{\threecolwid}
+\begin{beamercolorbox}[center,wd=\textwidth]{postercolumn}
+
+        \vfill
+
+        % -----------------------------------------------------------
+        % 3-1
+        % -----------------------------------------------------------
+%NOTE: results here need to be finessed in Mplus. will work on later.
+
+%         \begin{block}{Results, 20-50 years}
+% 
+% 
+% <<fit2>>=
+% @
+%  
+% % see http://tex.stackexchange.com/questions/94078/resize-table-and-caption
+% \centering
+% Model fit information
+% \vskip0.5ex
+% 
+% \begin{mdframed}[style=MyFrame]
+% \begin{minipage}[h]{1\textwidth}                          % minipage spans half the textwidth
+%         \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
+% <<printz-4, results='asis', eval=T>>=
+%         print(z,type="latex")
+%         @
+%         \end{adjustbox}
+% \end{minipage}
+% \end{mdframed}
+% \vskip1ex
+% 
+%         <<print-combo-race-50>>=
+%         @
+% 
+% \centering
+% Model 1 parameters for race/ethnic groups
+% 
+% \begin{mdframed}[style=MyFrame]
+% \begin{minipage}[h]{1.0\textwidth}                          % minipage spans half the textwidth
+%         \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
+% <<printz-5, results='asis', eval=T>>=
+%         print(z,type="latex")
+%         @
+%         \end{adjustbox}
+% \end{minipage}
+% \end{mdframed}
+% 
+%         \end{block}
+% 
+%         \vskip1ex
+%         \vfill
+
+        % -----------------------------------------------------------
+        % 3-2
+        % -----------------------------------------------------------
+\begin{block}{Results, 12-19 years, cont...}
+
+\centering\normalsize\textbf{Lipid construct comparisons across:}
+
+%\centering\normalsize \textbf{Across racial/ethnic groups} %see https://en.wikibooks.org/wiki/LaTeX/Colors
+  \begin{description}\normalsize
+                \item[Racial/ethnic groups] Evidence for metric non-invariance ($\chi^2$(df)=155 (8), p-value<0.001) $\Rightarrow$ Cannot make meaningful comparisons across racially and ethnically diverse populations.
+                \item[Time (years)] Evidence for metric non-invariance  ($\chi^2$(df)=155 (12), p-value<0.001)  $\Rightarrow$ Cannot make meaningful comparisons across time.
+      \end{description}
+\vskip0.5em
+% See examine-loadings-rev-chi-sq-tests.xlsx for Satorra-Bentler Scaled Chi-Square test (https://www.statmodel.com/chidiff.shtml)
+
 <<print-combo-race,  warning=TRUE>>=
@@
 
-\begin{frame}{Model 1 standardized parameters by race/ethnic groups}
+\centering
+\normalsize Model 1 standardized parameters by race/ethnic groups
 
-\centering
-\begin{minipage}[h]{0.9\textwidth}                          % minipage spans half the textwidth
+\begin{mdframed}[style=MyFrame]
+\begin{minipage}[h]{1.0\textwidth}                          % minipage spans half the textwidth
         \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
@@
 \end{minipage}
-\end{frame}
+\end{mdframed}
 
-\begin{frame}{Latent factor characteristics by racial/ethnic groups}
+\normalsize\centering\textbf{Latent factor characteristics:}
+  \begin{description}\normalsize
+            \item[Factor loadings] A one unit factor increase represents less favorable lipid value $\Rightarrow$ an increase in LDL-C (0.4) and TG (0.6) accompanied by a decrease in HDL ($\sim$ -0.6).
+            \item[Intercepts] Non-Hispanic Black group only one with positive HDL-C (0.132) and lowest TG (-1.078) relative to center of multivariate distribution.
+          \end{description}
+
+% Model fitting process
+%           \begin{itemize}
+%           \item Covariance between HDL-C and TG no longer significant when accounting for BMI covarying with latent lipid factor.
+%           \item Age positive covariance with lipid factor.
+%           \item BMI positive covariance with lipid factor.
+%           \end{itemize}
+
+        \end{block}
+        \vskip1ex
+
+        % -----------------------------------------------------------
+        % 3-3
+        % -----------------------------------------------------------
+        \begin{block}{Summary}
 
             \begin{itemize}
-            \item \underline{Factor loadings}: A one unit factor increase represents less favorable lipid value $\Rightarrow$ an increase in LDL-C (0.4) and TG (0.6) accompanied by a decrease in HDL ($\sim$ -0.6).
-            \item \underline{Intercepts}: Non-Hispanic Black group only one with positive HDL-C (0.132) and lowest TG (-1.078) relative to center of multivariate distribution.
+                \item[\Rightarrow] 3-indicator latent factor can function as a parsimonious and well-fitting measure for dyslipidemia in structural equation models.
+                \item[\Rightarrow] Cannot compare latent factor characteristics across groups by race/ethnic or time status.
             \end{itemize}
-\end{frame}
 
-<<print-combo-year,  warning=TRUE>>=
-@
+        \end{block}
+        \vskip1ex
+        \vfill
 
-\begin{frame}{Model 1 standardized parameters by year groups}
+        % -----------------------------------------------------------
+        % 3-3
+        % -----------------------------------------------------------
+        \begin{block}{References}
+          %\scriptsize{%          \printbibliography{}}
+          \scriptsize
+          (1) \fullcite{van_de_schoot_checklist_2012} \\
+          (2) \fullcite{brown_confirmatory_2015}
+          % see https://en.wikibooks.org/wiki/LaTeX/Fonts
+        \end{block}
 
-\centering
-\begin{minipage}[h]{0.8\textwidth}                          % minipage spans half the textwidth
-        \begin{adjustbox}{width=\linewidth}  % can also use \linewidth or sth. else
-<<printz-4, results='asis', eval=T>>=
-        print(z,type="latex")
-        @
-        \end{adjustbox}
-\end{minipage}
-\end{frame}
+\end{beamercolorbox}
+\end{column}
+% ---------------------------------------------------------%
+% end the 3rd column
+% ---------------------------------------------------------%
 
-% NOTE: not enough space here for gender. Remove to allow more
-% discussion of results in poster. keep for slides
-
-\begin{frame}{Fit and parameters for groups by sex}
-
-\vskip-1em
-
-% Run code for gender groups (from mplus_results.R)
-<<fit-gender>>=
-@
-\centering
-\begin{adjustbox}{width=0.7\linewidth} 
-%\caption*{Model fit}
-<<printf-gender, results='asis', eval=T>>=
-        print(z,type="latex",
-              caption="Model fit")
-@
-\end{adjustbox}
-
-\vskip1em
-
-<<print-combo-gender>>=
-@
-\centering
-\begin{adjustbox}{width=0.7\linewidth}
-%\caption*{Model 1 parameters}
-<<printp-gender, results='asis', eval=T>>=
-        print(z,type="latex",
-              caption="Model 1 parameters")
-@
-\end{adjustbox}
-
-\vskip1em
-
-\raggedright \footnotesize Invariance testing indicates evidence for metric non-invariance when testing difference between a metric model (model 1) and a full invariance model (model 4) in which there are full equality constraints across groups ($\chi^2$ = 68.5 (df=2); p-value<0.001).
-
-% NOTE: below is column approach, but too small to read
-
-% \begin{columns}[T] % align columns
-% \begin{column}{.48\textwidth}
-% 
-% %Left Part -------------------
-% 
-% % Run code for gender groups (from mplus_results.R)
-% <<fit-gender>>=
-% @
-% \centering
-% \begin{minipage}[h]{1\textwidth}  
-% \begin{adjustbox}{width=\linewidth} 
-% %\caption*{Model fit}
-% <<printf-gender, results='asis', eval=T>>=
-%         print(z,type="latex",
-%               caption="Model fit")
-% @
-% \end{adjustbox}
-% \end{minipage}
-% 
-% 
-% \end{column}%
-% 
-% \hfill%
-% \begin{column}{.48\textwidth}
-% 
-% 
-% % Right Part --------------------
-% 
-% <<print-combo-gender>>=
-% @
-% \centering
-% \begin{minipage}[h]{1\textwidth}  
-% \begin{adjustbox}{width=\linewidth}
-% %\caption*{Model 1 parameters}
-% <<printp-gender, results='asis', eval=T>>=
-%         print(z,type="latex",
-%               caption="Model 1 parameters")
-% @
-% \end{adjustbox}
-% \end{minipage}
-% 
-% \end{column}%
-% \end{columns}
+\end{columns}
 
 \end{frame}
-
-% >>>>>>>>>>>>>>>>>>>>>>>>>>>
-
-\begin{frame}{Model fitting process}
-          \begin{itemize}
-          \item Covariance between HDL-C and TG no longer significant when accounting for BMI covarying with latent lipid factor.
-          \item Age positive covariance with lipid factor.
-          \item BMI positive covariance with lipid factor.
-          \end{itemize}
-\end{frame}
-
-\section{Summary}
-\subsection{Summary}
-
-\begin{frame}{Summary} 
-\Large
-            \begin{itemize}
-                \item 3-indicator latent factor can function as a parsimonious and well-fitting measure for dyslipidemia in structural equation models.
-                \item []
-                \item Important not to compare latent factor characteristics across groups by race/ethnic, time or sex status given metric measurement non-invariance.
-            \end{itemize}
-\end{frame}
-
-\begin{frame}
-
-Contact information: vonholle at unc dot edu
-
-Further project information: https://avonholle.github.io/cfa-lipids/
-
-\end{frame}
-
 \end{document}

### 2017-03-06 StatisticalAnalysisPlan.Rmd
initial commit programs

---
title: "Statistical Analysis Plan (SAP): Multivariate lipid distribution differences across time and race/ethnic groups: United States, 2003-2012"
author: "Ann Von Holle"
date: '`r format(Sys.time(), "%d %B, %Y")`'
bibliography: lit.bib
output:
  html_document:
    toc: yes 
    toc_float: yes
    number_sections: true
---

<!-- NOTES:
1. This is the source document for this project. I will treat this document as a plain text Markdown file to create Word docs, pdfs and html files. This document will be downloaded in plain text format so any formatting such as bullet points, font changes, bolding, header choices will not be saved. However, please add comments or edits as you see fit. I can clean up later.

2. Any text below preceded by an @ symbol is a placeholder for a citation.

3. To indicate formatting in plain text use the following notation.
Sections in this document start with a # symbol. Section 1 headings start with #, section 2 headings start with ##, etc..
**bold**
*italic*
$latex formula$

4. If you want to see what the text below looks like after formatting, you can cut and paste in https://stackedit.io/editor OR you can go to the web page with a recent version of updates in html format: html format or Word .docx format.

5. Used google code at following link to create heading numbering: http://stackoverflow.com/questions/12389088/google-docs-drive-number-the-headings

6. Can edit this file at https://docs.google.com/document/d/1UI7c676DO4_ItDheW9PGy3yBsRTh1l_EPDOGlo3sIkA/edit?usp=sharing

END OF NOTES
-->

**Version: 1**
[Please click here for location of working draft in google docs](https://docs.google.com/document/d/1UI7c676DO4_ItDheW9PGy3yBsRTh1l_EPDOGlo3sIkA/edit?usp=sharing)

Note: Borrowed format of SAP from [Pfizer](https://www.pfizer.com/files/research/research_clinical_trials/Clinical_Data_Access_Request_Sample_SAP.pdf)

<!-- Statistical Analysis Plan
SAP should delineate populations (exclusion criteria); end points; descriptive objectives; testable hypotheses; modifications or derivations of standard variables; statistical methods, including handling of missing data, correlated data, bias, and confounding; subgroups; interactions; and sensitivity analysis.
-->

# Abstract

Plasma concentrations of low density lipoprotein cholesterol (LDL-C), high density lipoprotein cholesterol (HDL-C), and triglycerides (TG) are important risk factors for metabolic and cardiovascular disease c and are widely-used as targets for therapeutic intervention. A consideration of the multivariate distribution of these lipid traits could be informative in studies considering the impact of metabolic dysregulation on dyslipidemia and in studies of liver disease and the joint impact of dyslipidemia on insulin resistant states. Indeed, an understanding of the joint action of these traits has been relatively unexplored in the literature. Thus, the primary aim of this study is to determine the multivariate distribution of HDL-C, LDL-C and TG represented by racial/ethnic groups and stratified by age over five 2-year time periods (2003-2004, 2005-2006, 2007-2008, 2009-2010, 2011-2012) from NHANES. As these lipid values biologically related and statistically correlated, we first used principal component analyses (PCA) as an exploratory exercise to visually examine clustering by race in biplots. Second, we followed the PCA with tests of covariance and relative mean differences in the multivariate lipid distribution across a) calendar time and b) racial/ethnic groups (Mexican-American, non-Hispanic White, non-Hispanic Black, Other Hispanic and Other race/ethnicity). All analyses were stratified within three age groups: 12-19 years, 20 to 49 years and 50-80 years, and adjusted for age, race, sex and body mass index (BMI) when possible. Confirmatory factor analyses (CFA) allowed hypotheses tests stipulating covariance and mean relative location differences across groups using one latent factor/trait and HDL-C, LDL-C and TG serving as indicator variables for the factor. We found no evidence of covariance or relative location differences (alpha level of 0.05) across time but differences do exist across different racial/ethnic groups. In particular, covariance / relative locations differed across the five racial/ethnic groups for ages 12-19 years (p<0.0001 / p<0.0001), 20-49 years (p<0.02 / p<0.0001) and 50-80 years (p<0.003 / p<0.0001). These results suggest joint distributions of TG, HDL-C, and LDL-C vary across racial/ethnic groups. In summary, our data suggest that consideration of a multivariate distribution can provide a more nuanced portrait of dyslipidemia. One continuous latent factor representing all three lipids and their concomitant associations provides a means to characterize lipid values simultaneously and is a natural choice as an outcome when studying exposures influencing multiple lipid values, pleiotropic genetic variants being one case in point. We described distinct latent factors across racial/ethnic groups.  Limited analyses exist using this framework for lipids, and future efforts considering the joint distribution  of lipid measures may improve our understanding of the disease processes operation on metabolic dysregulation, for example in liver disease.

Character Maximum Limit 2500
Character Minimum Limit 25

# Data Source

[National Health and Nutrition Study](http://www.cdc.gov/nchs/nhanes/)
Years: 2005 to 2014

We will use the [nhanes-A R package](https://cran.r-project.org/web/packages/nhanesA/index.html) to obtain and download NHANES lab and demographic data sets.

Data sets for TG and LDL-C: TRIGLY_D, TRIGLY_E, TRIGLY_F, TRIGLY_G, TRIGLY_H
Data sets for HDL-C: HDL_D, HDL_E, HDL_F, HDL_G, HDL_H
Data sets for demographic info: DEMO_D, DEMO_E, DEMO_F, DEMO_G, DEMO_H

# Analysis Objectives

To explore the multivariate distribution of HDL-C, LDL-C and TG we will use principal component analysis (PCA) for complex survey samples (using appropriate weights and strata) to estimate proportion explained variance, factor loadings and create biplots to visualize data.

# Analysis Sets -- Populations -- Subgroups

## Subgroups

All analyses will be stratified by two-year groups: 2005-2006, 2007-2008, 2009-2010, 2011-2012, 2013-2014.

The data sets for 1999-2000, 2001-2002, 2003-2004 are available, but use a different measurement technique for HDL-C.

# Outcomes and Covariates

## Outcome variables

HDL-C (mg/dL): LBDHDD

LDL-C (mg/dL): LBDLDL

TG (mg/dL): LBXTR
Note: TG will be natural log transformed.

## Covariates

Age: RIDAGEYR
  Age groups will be 8-17 and 18-55

Gender: [RIAGENDR](http://wwwn.cdc.gov/Nchs/Nhanes/2013-2014/DEMO_H.htm#RIAGENDR)

Racial/Ethnic Group: [RIDRETH1](https://wwwn.cdc.gov/Nchs/Nhanes/2013-2014/DEMO_H.htm#RIDRETH1])

| --- | --- |
|Code | Value |
| 1 | Mexican American |
| 2 | Other Hispanic |
| 3 | Non-Hispanic White |
| 4 | Non-Hispanic Black |
| 5 | Other Race - Including Multi-Racial |
| . | Missing |
| --- | --- |

Note: In 2007-2008 NHANES over-sample all Hispanics not just Mexican- Americans as before.

### Cholesterol medication use

Create a binary variable indicating any use of cholesterol lowering medication. We used the definition of cholesterol-lowering medication use from the NCHS data brief 177 [@quiping_prescription_2014], “Prescription medications were classified based on the three level nested therapeutic classification scheme of Cerner Multum’s Lexicon (8). Cholesterol lowering medications were identified using the secondary level of drug category codes, specifically code 19. Another prescription drug, omega-3-polyunsaturated fatty acid (brand name Lovaza or Omacor), was also considered a cholesterol-lowering medication.”

[RXQ_DRUG data set](http://wwwn.cdc.gov/nchs/nhanes/1999-2000/RXQ_DRUG.htm)  contains the prescription medication info. If RXDDCI1B or RXDDCI2B or RXDDC3B or RXDDCI4B = 19 or RXDDRUG contains ‘OMEGA’ then treat as a cholesterol-lowering drug. Merge with drug data for individuals available at [rxq_rx](http://wwwn.cdc.gov/nchs/nhanes/1999-2000/RXQ_RX.htm), [rxq_rx_b](https://wwwn.cdc.gov/Nchs/Nhanes/2001-2002/RXQ_RX_B.htm) to [rxq_rx_g](http://wwwn.cdc.gov/nchs/nhanes/2011-2012/RXQ_RX_G.htm).

# Handling of missing values and other data conventions

If there are fewer than 10% missing data for lipid measures we will use complete cases. Otherwise I am not sure how to handle missing data with a complex survey sample and PCA.

# Statistical Methodology

## Statistical procedures

First, we will create summary statistics for all five time periods with weighted marginal and joint distributions by race/ethnicity, age and gender. Descriptive sample statistics will also include time-specific three dimensional plots for the three lipid measures with and without a legend identifying racial and ethnic groups.

We will use the three individual-level lipid measures to estimate proportion of variance explained for each of the three principal components. We will select the appropriate number of components to display in biplots based on the most variance explained. Biplots will be created for each year and age group combination (7 x 3 = 21 plots). In each of these plots we will use the race/ethnic groupings provided by NHANES as different colors to examine any groupings that may occur in the combined plot. We will use the biplot to examine the placement of each vector representing a lipid value for information on correlation between measures and strength of association.

All of the estimates mentioned above will be compared in tables (factor loadings) and plots (biplot),

Subsequent to this first analysis, a secondary analysis by year, age, and race/ethnic group will be completed to examine any difference in factor loadings and the relative importance of each lipid level within principal component (5 X 2 X 4).

NOTE: restricting these analyses to the four largest racial/ethnic groups as coded in the [NHANES sample](https://wwwn.cdc.gov/Nchs/Nhanes/2013-2014/DEMO_H.htm#RIDRETH1): 1) Mexican-American, 2) Non-Hispanic White, 3) Non-Hispanic Black and 4) Other Race.

### Secondary statistical procedures

Apply nonparametric hierarchical clustering methods (R package [hclust](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html)) to the coordinates in the first and second principal components. Start with number of clusters suggested in a mixture model (R package [mclust](https://cran.r-project.org/web/packages/mclust/vignettes/mclust.html)) via the BIC statistic.
## Measures to adjust for multiplicity, confounders, heterogeneity

These analyses are descriptive so no adjustments done for confounding. Analyses are stratified by age, time and racial/ethnic groups.

# Sensitivity analyses

When possible we will examine all plots stratified by gender to determine if there are differences.

# Rationale for any deviation from pre-specified analysis plan

If there are differences by gender in results we will stratify when the sample size is large enough.

#  QC plans

We will create histograms and descriptive summary statistics to examine data for any extreme values.

# Programming plans

All programs will be maintained/documented in a github repository [here](https://github.com/avonholle/ms201608-2). Note that this repository is private. To access the repository please provide me with a github id and I will add you to this repository. If you do not have an id you can sign up for a free account [here](https://help.github.com/articles/signing-up-for-a-new-github-account/).

We will use the ‘survey’ package in R to calculate principal component analyses (PCA). The **svyprcomp** function in this package accounts for the complex sample survey design in estimating scores.
Preliminary results can be found [here](https://avonholle.github.io/ms201608-2/).

# Appendix

# References

### 2017-03-06 mplus-check1.Rmd
initial commit programs

---
title: "Mplus: sensitivity analysis for useobservations vs subpopulation command"
author: "Ann Von Holle"
date: '`r format(Sys.time(), "%d %B, %Y")`'
bibliography: lit.bib
output:
  html_document:
        toc: yes
        toc_float: yes
        number_sections: true
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}

library(knitr)
opts_chunk$set(fig.width=10, fig.height=10, 
               fig.keep = 'all',
               fig.align="left", 
               echo=F, results='hide',
               warning=FALSE, message=FALSE, comment=NA)

library("MplusAutomation")
library(data.table)
```

<!-- Run mplusautomate package to extract results from mplus output, .out 
 See https://cran.r-project.org/web/packages/MplusAutomation/vignettes/Vignette.pdf
-->

```{r}
# run analyses with subpopulation (no grouping)

runModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/sensitivity', replaceOutfile="never")

```

```{r}
# read in parameter estimates from model using subpopulation and useobservatios options.

# extract out all parameters from output
sub.params <- extractModelParameters("C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/sensitivity", recursive=TRUE)

names(sub.params)
names(sub.params) = c("sub", "use") # rename to shorter name

```

```{r}
# Compare model parameters (and se). Look for differences.

names(sub.params$sub$unstandardized)
sub = data.frame(sub.params$sub$unstandardized)
use = data.frame(sub.params$use$unstandardized)

sub

# make a new value to put in table: est (se)
sub = within(sub, {
  est.se = paste0(est, " (", se, ")")
  option="sub"
})

use = within(use, {
  est.se = paste0(est, " (", se, ")")
  option="use"
})

tot = rbind(sub, use)

pct.diff.se = 100*((use$se - sub$se)/sub$se)

t1 = data.table(dcast(setDT(tot), paramHeader + param ~ option, 
                      value.var = "est.se"))

t1$pct.diff.se = round(pct.diff.se,2)
t1

colnames(t1) = c("parameter", "lipid indicator", "est(se) from sub", 
                 "est(se) from use", "% diff se")
```

# Table comparing parameter results using SUBPOPULATION option in complex design versus USEOBSERVATIONS option.

```{r, results='markup'}
  kable(t1)
```

Problem with complex design and GROUPING option: To determine measurement invariance I need to use the grouping option. However, with you can't use the SUBPOPULATION option and GROUPING command together. So I am checking to see if there is a big difference when using useobservations instead of subpopulation. Based on findings in table there is not a big difference. If anything the standard errors appear to be slightly larger if there is any chance at all.

### 2017-03-06 mplus-results.Rmd
initial commit programs

---
title: "Mplus Results"
author: "Ann Von Holle"
date: '`r format(Sys.time(), "%d %B, %Y")`'
bibliography: lit.bib
output:
  html_document:
        toc: yes
        toc_float: yes
        number_sections: true
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}

library(knitr)
opts_chunk$set(fig.width=10, fig.height=10, 
               fig.keep = 'all',
               fig.align="left", 
               echo=F, results='hide',
               warning=FALSE, message=FALSE, comment=NA)

library("MplusAutomation")
library(data.table)
library(plyr)

library(ztable)
options(ztable.type="html")
#options(ztable.type="viewer") # for debugging ztable

```

```{r, eval=FALSE}
# generate data to read in to Mplus
  read_chunk("export-mplus.R")
```

```{r}
# Look at some descriptive statistics in data

load("alldat.Rda") # get appended data (all.dat.df) from export-mplus.R
summary(all.dat.df)

# get parameters to set in mplus file templates for variance of bmi and age

# mean, sd bmi
with(all.dat.df, {
  c(mean(bmi, na.rm=T),
    sd(bmi, na.rm=T),
    mean(age, na.rm=T),
    sd(age, na.rm=T))})

table(all.dat.df$age.cut)

# get mean bmi and age by age cut groups
all.dt = data.table(all.dat.df[c("bmi", "age", "age.cut")])
all.dt[, sapply(.SD, function(x) list(mean=mean(x, na.rm=T), 
                                      sd=sd(x, na.rm=T))), 
       by=age.cut]

```

<!-- Make the models from an MplusAutomate template -->

```{r}

# Models comparing racial/ethnic groups and time (years)

# Each template iterates across age groups and groupings (year and race)
# terms from model come from "A checklist for testing measurement invariance",
# doi: 10.1080/17405629.2012.686740
# ----------------------------------------------------------------

# Template for Model 1: metric invariance. 
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m1.txt')

# Template for Model 2: Intercept only invariance
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m2.txt')

# Template for Model 3: Scalar invariance
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m3.txt')

# Template for Model 4: Full uniqueness invariance
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m4.txt')

# NOTE: after creating the models go to virtuallab and run them in MPlus. Details below.cc
```

```{r}

# Models comparing gender groups

# Each template iterates across age groups

# terms from model come from "A checklist for testing measurement invariance",
# doi: 10.1080/17405629.2012.686740
# ----------------------------------------------------------------

# Template for Model 1: metric invariance. 
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m1_sex.txt')

# Template for Model 2: Intercept only invariance
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m2_sex.txt')

# Template for Model 3: Scalar invariance
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m3_sex.txt')

# Template for Model 4: Full uniqueness invariance
createModels('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template/template_m4_sex.txt')

```

<!-- Run models created from template.
    Notes:
    1. I originally worked out all the problems in files for individual models/groups and age groups in folder titled, 
    ~\Dropbox\unc.grad.school\my-papers\ms-201608-2\programs-ms201608-2\mplus\archive\ 
    2. this won't run for the virtual lab Mplus. trying to figure out the command line to run mplus 
-->

```{r, eval=FALSE}

# -----------------------------------
# Note: I was able to run batches of Mplus files from the virtuallab.unc.edu server
# after emailing tech support 9/23/2016. A special version of R on virtuallab
# was set up for me to be able to call Mplus in virtuallab. was able to run
# all 28 models in less than 5 minutes after that.

# Contents of run-models.R program below (but note that it has to be run in R while in virtuallab.unc.edu). Before running run-models.R I run the chunk of code above to get the .inp files on my local folder (c:\temp\models\). then I copy those .inp files to the c:\temp\models folder. next I copy and paste the c:\temp\models\ folder on my local computer to my temporary folder in virtuallab.unc.edu server using their version of windows explorer and typing "%userprofile%". this pulls up my temporary server folder that R in virutal lab can locate. I also copy and paste the totdatlong.dat file made in 'export-mplus.R' (also in c:\temp\models\ folder) to the c:\users\vonholle\virtual-lab\ folder (or whatever you name it to in the virtual lab desktop, could be c:\users\vonholle\temp) to folder holding my programs in virtuallab.unc.edu. Then I run the run-models.R program and it should create .out files in the c:\users\vonholle\virtual-lab\ folder. Then I copy and paste those .out files back to my local drive (at ~\Dropbox\unc.grad.school\my-papers\ms-201608-2\programs-ms201608-2\mplus\compare-years-template) so I can run the rest of this .Rmd file.
# -----------------------------------

# install.packages("MplusAutomation")
# library(MplusAutomation)
# 
# runModels("C:\\Users\\vonholle\\temp")

```

<!-- Extract results from mplus output, .out 
 See https://cran.r-project.org/web/packages/MplusAutomation/vignettes/Vignette.pdf
-->

```{r read1, cache=F}

# read in results for differences across years (all ages) and racial/ethnic groups
# NOTE: I have to move the files from c:\temp\models\ to the folder specified below

# extract out all parameters from output
amp.years.params <- extractModelParameters("C:/temp/models2", recursive=F)
warnings() # not sure why there are some problems reading in data -- missing values?

#NOTE: be careful with the recursive setting. will get repeats that mess up tables below.

ayp = amp.years.params # save a backup because it takes so long to run
names.ayp = names(ayp)

# extract out model fit summaries from output
amp.years.sums <- extractModelSummaries("C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2/mplus/compare-years-template", recursive=T)
names(amp.years.sums)
```

```{r data-handle1}

# check names of objects in list
names(amp.years.params) # names too long, simplify

nms = names.ayp

gender.g = ifelse(grepl("sex.all.", nms)==T, 'all',
                  ifelse(grepl("sex.fem.", nms)==T, 'female',
                         ifelse(grepl("sex.mal.", nms)==T, 'male', 
                                ifelse(grepl("groups.gender", nms)==T, 'gen', NA))))
table(gender.g) # check

ageg = ifelse(grepl("12to19",nms)==T, "12to19",
              ifelse(grepl("20to50",nms)==T,"20to50",
                     ifelse(grepl("50to80",nms)==T, "50to80", NA)))

table(ageg) #check

gps = ifelse(grepl("groups.year",nms)==T, "YR",
             ifelse(grepl("groups.race", nms)==T, "RE",
                    ifelse(grepl("groups.gender", nms)==T, "GE", NA)))
table(gps)

model = ifelse(grepl(".m1", nms)==T, 1,
               ifelse(grepl(".m2", nms)==T, 2,
                      ifelse(grepl(".m3", nms)==T, 3,
                             ifelse(grepl(".m4", nms)==T, 4, NA))))
table(model)

names(amp.years.params) = paste0("m", model, "y", ageg, "g", gps, "gender", gender.g)
names(amp.years.params) #check
head(amp.years.params)
```

```{r data-handle2}
# --------------------------------------------
# PREP model fit info
# --------------------------------------------

head(amp.years.sums)

# add descriptors to model fit data

amp.years.sums$gender.g = with(amp.years.sums, {
  ifelse(grepl("-all-", Filename)==T, 'all',
                  ifelse(grepl("-female-", Filename)==T, 'female',
                         ifelse(grepl("-male-", Filename)==T, 'male', NA)))
})

table(amp.years.sums$gender.g) # check

amp.years.sums$agegrp = with(amp.years.sums, {
  ifelse(grepl("12to19",Filename)==T, "12to19",
                                 ifelse(grepl("20to50",Filename)==T,"20to50",
                                        ifelse(grepl("50to80",Filename)==T, "50to80", NA)))
})
table(amp.years.sums$agegrp)

amp.years.sums$group = with(amp.years.sums, {
  ifelse(grepl("groups.year",nms)==T, "YR",
             ifelse(grepl("groups.race", nms)==T, "RE",
                    ifelse(grepl("groups.gender", nms)==T, "GE", NA)))
})

table(amp.years.sums$group)
names(amp.years.sums)

amp.years.sums$model = with(amp.years.sums, {
  ifelse(grepl("m1-", Filename)==T, 1,
               ifelse(grepl("m2-", Filename)==T, 2,
                      ifelse(grepl("m3-", Filename)==T, 3,
                             ifelse(grepl("m4-", Filename)==T, 4, NA))))
})

table(amp.years.sums$model)

# Get columns to print off table below
names(amp.years.sums)
head(amp.years.sums)
fit.vals = c("Filename",
             "ChiSqM_Value", "ChiSqM_DF", "ChiSqM_PValue", 
             "LL",
             "CFI", "TLI", "RMSEA_Estimate",
             "BIC", "AIC", "gender.g", "agegrp", "group", "model")

fit.vals.renamed = c("Chi-sq", "df", "p-value", "LL",
                     "CFI", "TLI",
                     "BIC", "AIC", "RMSEA", "Model", "Gender", "Age", "Groups", "mod")

amp.years.sums.sub = amp.years.sums[,colnames(amp.years.sums) %in% fit.vals]
colnames(amp.years.sums.sub) = fit.vals.renamed

head(amp.years.sums.sub) #check

# make a model description to describe model invariance settings as described in
# van de Schoot R, Lugtig P, Hox J. A checklist for testing measurement invariance. European Journal of Developmental Psychology. 2012 Jul;9(4):486–92.

amp.years.sums.sub = within(amp.years.sums.sub, {
  description = ifelse(grepl("m1", Model)==T, "M1: Metric invariance: diff int",
                       ifelse(grepl("m2", Model)==T, "M2: Diff factor loadings",
                              ifelse(grepl("m3", Model)==T, "M3: Scalar Invariance: eq loading and eq int",
                                     ifelse(grepl("m4", Model)==T, "M4: Full Uniqueness invariance: eq loadings, intercepts and res variance",
                                            ifelse(grepl("m5", Model)==T, "M5: Free loadings, int and res variance",
                                                   NA)))))
})

table(amp.years.sums.sub$Groups)

order = amp.years.sums.sub[order(amp.years.sums.sub$Age,
                         amp.years.sums.sub$Groups,
                         amp.years.sums.sub$description),]

  mod.1 = order[order$Gender=="all" & order$Groups %in% c("RE", "YR"), !(colnames(order) %in% c("Model"))]
  table(mod.1$Groups)
  head(mod.1)

  mod.1f = order[order$Gender=="female"& order$Groups %in% c("RE", "YR"),
                 !(colnames(order) %in% c("Model"))]
  head(mod.1f)
  table(mod.1f$Groups)

  mod.1m = order[order$Gender=="male"& order$Groups %in% c("RE", "YR"),
                 !(colnames(order) %in% c("Model"))]
  head(mod.1m)

  mod.ge = order[order$Groups %in% c("GE"), !(colnames(order) %in% c("Model"))]
  head(mod.ge)

  mod.1.tot = rbind(mod.1[complete.cases(mod.1$df),], 
                    mod.ge) #includes gender groups (in addition to race/ethnic and year groups)

  table(mod.1.tot$Groups)
  mod.1.tot = mod.1.tot[order(mod.1.tot$Groups, mod.1.tot$Age, mod.1.tot$description),]
  mod.1.tot[1:10,]

```

# Model fit results for pooled gender

Note: All factors were adjusted for age and bmi (in the year comparisons).

```{r fit1}

  col.include = c("mod", "Chi-sq", "df", "p-value",
                  "LL", "CFI", "TLI", "BIC", "AIC", "RMSEA",
                  "description")
  mod.2 = mod.1.tot[col.include]
  mod.2$mod = as.integer(mod.1.tot$mod)
  colnames(mod.2)[1] = "Model"
  head(mod.2)

  z = ztable(mod.2, zebra=4, zebra.color="pastelgray")

  # Add row labels
  rowlabs = with(mod.1.tot, {
    unique(paste0("Age: ", Age, ". Groups:", Groups))
  })
  rowlabs

    rowlabs.n = with(mod.1.tot, {
    table(paste0("Age: ", Age, ". Groups:", Groups))
  })

  z = addrgroup(z, rgroup = rowlabs, n.rgroup = rowlabs.n)
  z = update_ztable(z, include.rownames=F, 
                    caption="Model Fit Information")
```

```{r, results='asis'}
z

```

# Test model fit

Use nested chi-square tests to compare models 1 to 3 (some form of variance across groups) to model 4(complete invariance across groups) [@van_de_schoot_checklist_2012]

* **Model 1**. metric invariance: Factor loadings are eq across groups but int differ across groups (metric invariance)
* **Model 2**. Intercept only invariance: factor loadings are different but interceptes are eq across groups
* **Model 3**. Scalar invariance: Same factor loadings and same intercepts for underlying items
* **Model 4**. Full uniqueness invariance: equal loadings and intercept. Also equal variance across all groups

Separate models were run for groupings: RE = racial/ethnic groups, YE = years (2003-2004, 2005-2006, 2007-2008, 2009-2010, 2011-2012)

```{r}

# reshape data with chi-sq values for each model in columns by types of model (age and grouping combos)
head(order)

get.pvals = function(df1) {
  #df1=mod.1.tot
  df1 = df1[complete.cases(df1$df),]
  table(df1$Groups)
  nrow(df1)
  head(df1)
  df1

  chi = dcast(setDT(df1), Age + Groups ~ mod, 
                        value.var = "Chi-sq")

  colnames(chi)[3:6] = c("m1", "m2", "m3", "m4")
  chi

  df = dcast(setDT(df1), Age + Groups ~ mod, 
                        value.var = "df")

  colnames(df)[3:6] = c("m1", "m2", "m3", "m4")
  df

  # get diff in chi sq for model 4 vs 3, 2 and 1
  chi = within(chi, {
    diff.4.vs.1 = m4-m1
    diff.4.vs.2 = m4-m2
    diff.4.vs.3 = m3-m4
  })

  chi

  # get diff in chi sq for model 4 vs 3, 2 and 1
  df = within(df, {
    df.4.vs.1 = m4-m1
    df.4.vs.2 = m4-m2
    df.4.vs.3 = m3-m4
  })

  df = data.frame(df)
  chi = data.frame(chi)

  both.vals = cbind(chi[c("Age", "Groups",
                          "diff.4.vs.1",
                          "diff.4.vs.2",
                          "diff.4.vs.3")],
                    df[c("df.4.vs.1",
                         "df.4.vs.2",
                         "df.4.vs.3")])

  both.vals

  # These models are nested so use difference in chi-sq tests to determine if 
  # any changes are statistically different from the full uniqueness invariance model (model 4)
  both.vals = within(both.vals, {
    pval.m1 = round(pchisq(diff.4.vs.1, df.4.vs.1, lower.tail=F),4)
    pval.m2 = round(pchisq(diff.4.vs.2, df.4.vs.2, lower.tail=F),4)
    pval.m3 = round(pchisq(diff.4.vs.3, df.4.vs.3, lower.tail=F),4)
  })

  return(both.vals)
}

bv = get.pvals(mod.1.tot)

bv.m = get.pvals(mod.1m)
bv.f = get.pvals(mod.1f)

```

```{r}

make.z.tests = function(df.vals, type){
  #df.vals is the data frame
  #type is the type of labeling for the final table

  cgroups = c( "", 
                          "Chi sq diff",
                          "df",
                          "p-value"
                          )

  both.vals.2 = df.vals[, !(colnames(df.vals) %in% c("Age"))]
  colnames(both.vals.2)
  both.vals.2 = both.vals.2[c("Groups", "diff.4.vs.1", "diff.4.vs.2", "diff.4.vs.3",
                              "df.4.vs.1", "df.4.vs.2", "df.4.vs.3",
                              "pval.m1", "pval.m2", "pval.m3")] # reorder order of columns
  colnames(both.vals.2) # check

  colnames(both.vals.2) = c("Groups", rep(c("Model 1 vs 4", "Model 2 vs 4", "Model 3 vs 4"),3))

  n.cgroup = c(1,3,3,3)
  z=ztable(both.vals.2, zebra=0, zebra.color="pastelgray")
  z = addcgroup(z, cgroup=cgroups, n.cgroup=n.cgroup)

  rgroup = c("12 to 19 years", "20 to 50 years", "50 to 80 years")
  if(type==1) {n.rgroup = c(rep(3,3))
                  z = addrgroup(z, rgroup = rgroup, n.rgroup = n.rgroup)}

  if(type==2) {n.rgroup2 = c(rep(2,3))
                z = addrgroup(z, rgroup = rgroup, n.rgroup = n.rgroup2)
  }

  z = update_ztable(z, include.rownames=F,
                    caption="Model fit tests",
                    zebra=2,
                    digits = c(0,0,rep(1,3), rep(0,3), rep(4,3))); #z
}

z.all = make.z.tests(bv, type=1)
z.f = make.z.tests(bv.f, type=2)
z.m = make.z.tests(bv.m, type=2)

```

## Both genders

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
  z.all
```

## Females

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
  z.f
```

## Males

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
  z.m
```

# Model parameters for all models

## Standardized factor loadings by age group, grouping (RE=race/ethnicity and YE=years) and model

```{r}

# YEARS
names(amp.years.params)

# pick off the standardized estimates, stdyx.standardized (3rd element in each list)
# and put in table for all models
names(amp.years.params)  # amp.years.params is a list of lists containing parameters from many different parts of the output for each model, indexed by model.
names(amp.years.params[[1]]$stdyx.standardized) # 7 columns
nrow(amp.years.params[[1]]$stdyx.standardized) # 130 rows
amp.years.params[[1]]$stdyx.standardized

# take out the standardized analysis from each of the model output (3rd element from each list, stdyx.standardized)
#res3 = do.call(rbind.data.frame, sapply(amp.years.params, function(x) x[3]))
#res3$names = rownames(res3)

res3 = rbindlist(sapply(amp.years.params, function(x) x[3]), 
                 use.names=TRUE, fill=TRUE, idcol="names")

head(res3)

# extract out model info (age and group)
#age.group = unlist(strsplit(rownames(res3), "[.]"))
res3$model = substr(res3$names,1,2)
res3$age.group = substr(res3$names,4,9)
res3$grouping = substr(res3$names,11,12)
res3$gender = substr(res3$names, 19, 21)

table(res3$model); table(res3$age.group); table(res3$grouping);
table(res3$gender)
head(res3[res3$Group=="FEMALE",])

# make a new value to put in table: est (se)
res3 = within(res3, {
  est.se = paste0(est, " (", se, ")")
})
head(res3$est.se)

# Look at factor loadings by model

fl = res3[res3$paramHeader %in% c("F1.BY"),]
nrow(fl) # 1065 rows
head(fl)

t1 = data.frame(dcast(setDT(fl[fl$gender %in% c("all", "gen"),]), 
                      age.group + grouping + model + Group ~ param, 
                      value.var = "est.se"))

colnames(t1)[1:4] = c("Age", "Grouping", "Model", "Group value")
head(t1)
table(t1$"Group value")

# females
t1f = data.frame(dcast(setDT(fl[fl$gender=="fem",]), age.group + grouping + model + Group ~ param, 
                      value.var = "est.se"))
colnames(t1f)[1:4] = c("Age", "Grouping", "Model", "Group value")
head(t1f)

# males
t1m = data.frame(dcast(setDT(fl[fl$gender=="mal",]), age.group + grouping + model + Group ~ param, 
                      value.var = "est.se"))
colnames(t1m)[1:4] = c("Age", "Grouping", "Model", "Group value")
head(t1m)

```

```{r}
# function to make standardized estimates in a z table

make.stdest = function(dat){
#dat=t1
  col.include = c("Group value", "HDL2", "LDL2", "TG2")
  col.names = c("Group", "HDL", "LDL", "TG")
  head(t1)
  dat.2 = dat[col.include]
  colnames(dat.2) = col.names
  head(dat.2)

  z = ztable(dat.2, zebra=0, zebra.color="pastelgray")

  # Add row labels
  rowlabs = with(dat, {
    unique(paste0("Age: ", Age, ". Groups:", Grouping, ". Model: ", substr(Model,2,3)))
  })
  rowlabs

  rowlabs.n = with(dat, {
    table(paste0("Age: ", Age, ". Groups:", Grouping, ". Model: ", substr(Model,2,3)))
  })

  z = addrgroup(z, rgroup = rowlabs, n.rgroup = rowlabs.n)
  z = update_ztable(z, include.rownames=F, 
                    caption="Factor Loadings (se)", 
                    zebra=2)
  return(z)
}

z = make.stdest(t1)
zm = make.stdest(t1m)
zf = make.stdest(t1f)

#  z
```

## Pooled gender

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
z
```

## Females

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
zf
```

## Males

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
zm
```

-------

## Standardized residual variance by age group, grouping (RE=race/ethnicity and YE=years) and model

```{r}
resid = res3[res3$paramHeader %in% c("Residual.Variances"),]
nrow(resid)
head(resid)

t2 = data.frame(dcast(setDT(resid[resid$gender %in% c("all", "gen"),]), 
                      age.group + grouping + model + Group ~ param, 
                      value.var = "est.se"))
colnames(t2)[1:4] = c("Age", "Grouping", "Model", "Group value")
t2

# females
t2f = data.frame(dcast(setDT(resid[resid$gender=="fem",]), age.group + grouping + model + Group ~ param, 
                      value.var = "est.se"))
colnames(t2f)[1:4] = c("Age", "Grouping", "Model", "Group value")

# males
t2m = data.frame(dcast(setDT(resid[resid$gender=="mal",]), age.group + grouping + model + Group ~ param, 
                      value.var = "est.se"))
colnames(t2m)[1:4] = c("Age", "Grouping", "Model", "Group value")

```

```{r}

# function to make z table of standardized residuals
makestdres = function(dat) {
  #dat=t2
  head(dat)
  col.include = c("Grouping", "HDL2", "LDL2", "TG2")
  col.names = c("Group", "HDL", "LDL", "TG")
  dat.2 = dat[col.include]
  colnames(dat.2) = col.names
  head(dat.2)
  names(dat)

  z = ztable(dat.2,  zebra=0, zebra.color="pastelgray")

  # Add row labels
  rowlabs = with(dat, {
    unique(paste0("Age: ", Age, ". Groups:", Grouping, ". Model: ", substr(Model,2,3)))
  })
  rowlabs

    rowlabs.n = with(dat, {
    table(paste0("Age: ", Age, ". Groups:", Grouping, ". Model: ", substr(Model,2,3)))
  })

  rgroup = rowlabs

  z = addrgroup(z, rgroup=rowlabs, n.rgroup = rowlabs.n)
  z = update_ztable(z, include.rownames=F, 
                    caption="Standardized residual variance (se)",
                    zebra=2)
}

z = makestdres(t2)
z

zf = makestdres(t2f)
zm = makestdres(t2m)
#  z
```

## Pooled gender

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
z
```

## Females

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
zf
```

## Males

**NOTE**: 
* GE = gender groups
* RE = racial/ethnic groups
* YR = year groups

```{r, results='asis'}
zm
```
----

## Standardized intercepts by age group, grouping (RE=race/ethnicity and YE=years) and model

```{r}
int = res3[res3$paramHeader %in% c("Intercepts"),]
nrow(int)
head(int)

t3 = data.frame(dcast(setDT(int[int$gender %in% c("all", "gen"),]), age.group + grouping + model + Group  ~ param, 
                      value.var = "est.se"))
colnames(t3)[1:4] = c("Age", "Grouping", "Model", "Group value")
head(t3)

t3f = data.frame(dcast(setDT(int[int$gender=="fem",]), age.group + grouping + model + Group  ~ param, 
                      value.var = "est.se"))
colnames(t3f)[1:4] = c("Age", "Grouping", "Model", "Group value")
head(t3f)

t3m = data.frame(dcast(setDT(int[int$gender=="mal",]), age.group + grouping + model + Group  ~ param, 
                      value.var = "est.se"))
colnames(t3m)[1:4] = c("Age", "Grouping", "Model", "Group value")
head(t3m)

```

```{r}
# function to show standardized intercept

makeint = function(dat){

  #dat=t3
  col.include = c("Group value", "HDL2", "LDL2", "TG2")
  col.names = c("Group", "HDL", "LDL", "TG")
  dat.2 = dat[col.include]
  colnames(dat.2) = col.names
  head(dat.2)

  z = ztable(dat.2, zebra=0, zebra.color="pastelgray")

  # Add row labels
  rowlabs = with(dat, {
    unique(paste0("Age: ", Age, ". Groups:", Grouping, ". Model: ", substr(Model,2,3)))
  })
  rowlabs

    rowlabs.n = with(dat, {
    table(paste0("Age: ", Age, ". Groups:", Grouping, ". Model: ", substr(Model,2,3)))
  })
  rowlabs.n

  z = addrgroup(z, rgroup=rowlabs, n.rgroup = rowlabs.n)
  z = update_ztable(z, include.rownames=F, 
                    caption="Standardized intercepts (se)", 
                    zebra=2)
}

z = makeint(t3)
zf = makeint(t3f)
zm = makeint(t3m)
#  z
```

## Pooled gender

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
z
```

## Females

**NOTE**: 
* GE = gender groups
* RE = racial/ethnic groups
* YR = year groups

```{r, results='asis'}
zf
```

## Males

**NOTE**:

* GE = gender groups

* RE = racial/ethnic groups

* YR = year groups

```{r, results='asis'}
zm
```

# References

### 2017-03-06 export-mplus.R

@@ -1,0 +1,280 @@
+# export-mplus.R
+
+# export nhanes data for mplus use (txt file)
+library(foreign)
+library("MplusAutomation")
+library(survey)
+library(reshape2)
+library(data.table)
+library(ggplot2)
+
+setwd('C:/Users/vonholle/Dropbox/unc.grad.school/my-papers/ms-201608-2/programs-ms201608-2')
+load("nhanes-dat.RData") # get data from read-data.R. Takes a while to run
+
+# Look at a small test data set firs
+# -------------------------------------------
+
+head(d.2005)
+names(d.2005)
+#d.2005$RIDAGEYR
+
+
+dat.2005 = subset(d.2005, chol.med==0 & RIDAGEYR>50, select=c(SEQN,
+                                                LBXTR, LBDLDL, LBDHDD,
+                                                RIDRETH1,
+                                                WTMEC2YR, SDMVPSU, SDMVSTRA))
+#table(dat.2005$SDMVPSU)
+#table(dat.2005$SDMVSTRA)
+head(dat.2005)
+
+# format some variables
+
+dat.2005$race = ifelse(dat.2005$RIDRETH1=="Non-Hispanic White", 1, 0)
+colnames(dat.2005) = tolower(colnames(dat.2005))
+#dat.2005$lbxtr = log(dat.2005$lbxtr)
+head(dat.2005)
+nrow(dat.2005)
+
+head(dat.2005)
+lipvars = c("lbxtr", "lbdldl", "lbdhdd")
+d2005m = dat.2005[,colnames(dat.2005) %in% lipvars]
+head(d2005m)
+cor(d2005m,  use="pairwise.complete.obs")
+
+dat.2005 = within(dat.2005, { # revise the lipid vars
+  lbxtr = scale(log(lbxtr)) # center and standardize log of tg
+  lbdldl = scale(lbdldl)
+  lbdhdd = scale(lbdhdd) # scale the inverse of hdl so it's in same direction as others in terms of adverse effect (positive assn with adverse risk)
+})
+
+
+summary(dat.2005)
+
+colnames(dat.2005)[colnames(dat.2005) %in% lipvars] = c("tg", "ldl", "hdl")
+
+head(dat.2005)
+
+prepareMplusData(dat.2005, "c:/temp/d2005.dat") # export data out to an mplus file
+
+
+# do a simple check with Mplus results for a simple linear regression with
+# ldl regressed on hdl
+
[…]

### 2017-03-06 mplus_results.R

@@ -1,0 +1,559 @@
+# mplus_results.R
+
+# All code borrowed from file to create html, mplus-results.Rmd
+# >>>>---------------------------------------------------------
+# Extract results from mplus output, .out 
+# See https://cran.r-project.org/web/packages/MplusAutomation/vignettes/Vignette.pdf
+
+library(knitr)
+library(data.table)
+library(readxl)
+library(ggplot2)
+library(plyr)
+library("MplusAutomation")
+library(ztable)
+options(ztable.type="viewer")
+library(htmlTable)
+
+
+# >>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
+## @knitr read1
+# >>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
+
+# PUT into .Rda files. takes too long to run
+# each time.
+
+# read in results for differences across years (all ages) and racial/ethnic groups
+# All data made in mplus-results.Rmd
+
+# extract out all parameters from output
+amp.years.params <- extractModelParameters("C:\\temp\\models2", 
+                                           recursive=F)
+warnings() # not sure why there are some problems reading in data -- missing values?
+names(amp.years.params)
+save(amp.years.params, file="201703-aha-poster-2\\ayp.Rda")
+
+#NOTE: be careful with the recursive setting. will get repeats that mess up tables below.
+# 
+# ayp = amp.years.params # save a backup because it takes so long to run
+# names.ayp = names(ayp)
+
+# extract out model fit summaries from output
+amp.years.sums <- extractModelSummaries("C:\\temp\\models2", recursive=F)
+names(amp.years.sums)
+save(amp.years.sums, file="201703-aha-poster-2\\ays.Rda")
+
+
+# >>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
+## @knitr data-handle1
+# >>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
+
+#rm(list=ls())
+
+#load("201703-aha-poster-2\\ayp.Rda") # for debugging
+load("ayp.Rda")
+# check names of objects in list
+#names(amp.years.params) # names too long, simplify
+nms = names(amp.years.params) 
+
+gender.g = ifelse(grepl("sex.all.", nms)==T, 'all',
[…]

### 2017-03-06 read-data.R

@@ -1,0 +1,314 @@
+# read-data.R ----------------------------------------------
+# -----------------------------------------------------------
+
+# example from the survey package documentation
+# ---------------------------------------------
+
+library(survey)
+library(Hmisc)
+library(nhanesA) # see https://cran.r-project.org/web/packages/nhanesA/vignettes/Introducing_nhanesA.html
+library(rgl)
+library(devtools)
+library(gridExtra)
+
+#install_github("vqv/ggbiplot")
+library(ggbiplot)
+
+library(data.table)
+
+# ----------------------------------------------------------------------------
+# following adapted from vignette for nhanesA: https://cran.r-project.org/web/packages/nhanesA/vignettes/Introducing_nhanesA.html
+# ----------------------------------------------------------------
+
+# Use nhanesA package to read in NHANES data from web page
+# --------------------------------------------------------
+
+# Download demographic data for merge
+
+demo = nhanes("DEMO") # http://wwwn.cdc.gov/Nchs/Nhanes/1999-2000/DEMO.htm
+demo_b = nhanes("DEMO_B") # http://wwwn.cdc.gov/Nchs/Nhanes/2001-2002/DEMO_B.htm
+demo_c = nhanes("DEMO_C") # http://wwwn.cdc.gov/Nchs/Nhanes/2003-2004/DEMO_C.htm
+demo_d = nhanes("DEMO_D") # http://wwwn.cdc.gov/Nchs/Nhanes/2005-2006/DEMO_D.htm
+demo_e = nhanes("DEMO_E") # http://wwwn.cdc.gov/Nchs/Nhanes/2007-2008/DEMO_E.htm
+demo_f = nhanes("DEMO_F")
+demo_g = nhanes("DEMO_G")
+demo_h = nhanes("DEMO_H")
+
+# Get drug info -------------------------------
+# -----------------------------------------------
+nhanesTables('Q', 1999)
+
+drug.info = nhanesSearch("Drug", ignore.case = T, ystart=1999, ystop=2014, data_group=c("QUESTIONNAIRE"))
+drug.info # this is not very informative
+
+rxq_rx = nhanes('RXQ_RX') # 1999-2000
+rxq_rx_b = nhanes('RXQ_RX_B') # 2001-2002
+rxq_rx_c = nhanes('RXQ_RX_C') # 2003-2004
+rxq_rx_d = nhanes('RXQ_RX_D') # 2005-2006
+rxq_rx_e = nhanes('RXQ_RX_E') # 2007-2008
[…]

### 2017-03-06 run-models.R
file to run Mplus programs

@@ -1,0 +1,6 @@
+# run-models.R
+
+install.packages("MplusAutomation")
+library(MplusAutomation)
+
+runModels("C:\\Users\\vonholle\\temp") 

### 2017-04-26 StatisticalAnalysisPlan.Rmd
updated sap

@@
 
+### Summary statistics
+
 First, we will create summary statistics for all five time periods with weighted marginal and joint distributions by race/ethnicity, age and gender.
@@
 
-We will use the three individual-level lipid measures to estimate proportion of variance explained for each of the three principal components.
-We will select the appropriate number of components to display in biplots based on the most variance explained.
-Biplots will be created for each year and age group combination (7 x 3 = 21 plots).
-In each of these plots we will use the race/ethnic groupings provided by NHANES as different colors to examine any groupings that may occur in the combined plot.
-We will use the biplot to examine the placement of each vector representing a lipid value for information on correlation between measures and strength of association.
+### Confirmatory Factor Analysis (CFA)
 
-All of the estimates mentioned above will be compared in tables (factor loadings) and plots (biplot),
+The primary analysis is a confirmatory factor analysis. This analysis includes the three most common lipid profile measures as a construct of dyslipidemia: HDL, LDL and TG.
 
-Subsequent to this first analysis, a secondary analysis by year, age, and race/ethnic group will be completed to examine any difference in factor loadings and the relative importance of each lipid level within principal component (5 X 2 X 4).
+<img src=cfa-figure.png width="300">
+<!-- This figure was made in cfa-figure.Rnw (and exported to .png from Adobe Acrobat DC) -->
+
+\[
+\text{Equations for CFA:}
+\left\{
+\begin{array}{rcl}
+  \textrm{HDL-C} & = & \mu_1 + \lambda_1 \eta_1 + \epsilon_1 \\
+  \textrm{LDL-C} & = & \mu_2 + \lambda_2 \eta_1 + \epsilon_2 \\
+  \textrm{TG-C} & = & \mu_3 + \lambda_3 \eta_1 + \epsilon_3 \\
+\end{array}
+\right.
+\]
+
+* Fit
+To evaluate fit of all models, we will consider:
+1. Root mean square error of approximation (RMSEA)
+2. Comparitive fit index (CFI)
+3. Bayesian Information Criterion (BIC)
+
+* Groups
+The two demographic groupings we will consider are:
+1. Racial/ethnic groups as defined in NHANES.
+2. Gender groups
+
+* Estimates
+Each group will have estimates for 
+1. Factor loadings, $\lambda$
+2. Intercept, $\mu$
+3. Residual variation, $\epsilon$
+
+---
+
+## Measurement Invariance (MI)
+
+To compare this construct across groups, measurement invariance (MI) of the construct must hold.
+To check MI, we will conduct a sequence of model fitting [@van_de_schoot_checklist_2012].
+If MI does not hold then the construct cannot be meaningfully compared across groups, such as race/ethnicity or gender.
+
+As defined in the checklist in the paper cited above by @van_de_schoot_checklist_2012, four models are fit with the same data including ones to evaluate:
+
+1. Metric invariance: test difference of factor loadings, $\lambda$.
+
+2. Equality of intercepts: test difference of intercepts, $\mu$.
+
+3. Scalar invariance: test equality of factor loadings, $\lambda$, and intercepts, $\mu$.
+
+4. Full uniqueness MI: “the explained variance is the same for every item is the same across groups.” [@van_de_schoot_checklist_2012]
 
@@
 
-### Secondary statistical procedures
-
-Apply nonparametric hierarchical clustering methods (R package [hclust](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html)) to the coordinates in the first and second principal components.
-Start with number of clusters suggested in a mixture model (R package [mclust](https://cran.r-project.org/web/packages/mclust/vignettes/mclust.html)) via the BIC statistic.
 ## Measures to adjust for multiplicity, confounders, heterogeneity
@@
 
-All programs will be maintained/documented in a github repository [here](https://github.com/avonholle/ms201608-2).
-Note that this repository is private.
-To access the repository please provide me with a github id and I will add you to this repository.
-If you do not have an id you can sign up for a free account [here](https://help.github.com/articles/signing-up-for-a-new-github-account/).
-
-We will use the ‘survey’ package in R to calculate principal component analyses (PCA). The **svyprcomp** function in this package accounts for the complex sample survey design in estimating scores.
-Preliminary results can be found [here](https://avonholle.github.io/ms201608-2/).
+All programs will be maintained/documented in a github repository [here](https://github.com/avonholle/ms-cfa-lipids).
 

### 2017-05-24 manuscript.Rmd
initial commit for manuscript

---
title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups: United States, 2003 to 2012"
author: "Ann Von Holle, Kari E. North"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true # table of content true
    depth: 3  # up to three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
bibliography: lit.bib
csl: ../../vancouver.csl
---

<!-- 
NOTES:

1. This document will be downloaded in text format so any formatting such as bullet points, font changes, bolding, header choices will not be saved. However, please add comments or edits as you see fit. I can clean up later.
2. Any text below preceded by an @ symbol is a placeholder for a citation.
3. To indicate formatting in plain text use the following notation.
Sections in this document start with a # symbol. Section 1 headings start with #, section 2 headings start with ##, etc..
**bold**
*italic*
$latex formula$
4. If you want to see what the text below looks like after formatting, you can cut and paste in https://stackedit.io/editor OR you can go to the web page with a recent version of updates in Word doc format: draft
END OF NOTES
-->

# Introduction

Dyslipidemia is often characterized through measures of triglycerides, high-density lipoprotein cholesterol (HDL-C), and low-density cholesterol (LDL-C) from a common lipid panel. These measures are then analyzed in research on a univariate basis or with ratios. Another analytic approach can include confirmatory factor analysis (CFA). In practice, this approach would reduce the dimension of three indicators mentioned above to one latent variable capturing the pairwise correlation between measures. In turn, the resulting latent construct can be used either as an exposure or outcome in structural equation models. Although this sort of multivariate analysis can provide a different perspective compared to univariate and traditional approaches, it has not been investigated in the literature as a feasible analytic approach.

CFA is useful for a set of indicators that are interrelated and have already been established in a substantive theoretical sense independent of a statistical rationale. For the three indicators of interest in this particular study: HDL-C, LDL-C, and TG, there exists a strong rationale to treat them together to represent dyslipidemia [@bays_national_2016]. Although there are other indicators of dyslipidemia, these three are the most accessible in common lipid profiles and the most reported in the literature. Any changes in these values represented by a latent factor would be recognizable to lipid researchers and of value in research. What would most likely occur is for TG and LDL-C to be positively correlated and HDL-C negatively correlated with the prior two indicators. A one unit change in the latent factor in this case could represent an increase in TG and LDL-C alongside a decrease in HDL-C given the well established correlations between these measures.

A further motivation to examine dyslipidemia in a dimension reduction approach is to further develop the rationale behind the use of ratios when characterizing lipids. One could consider the correlations between the indicators to be an analog to ratios, with an additional ability to maintain variability for each measure. For example, the use of HDL to TG ratios is increasing in the literature, but analyses using a ratio has limitations [@kuh_correlation_1955; @atchley_ratios_1978; @kronmal_spurious_1993] including failure to adequately adjust for each of the variables forming the ratio when used as a covariate in a linear regression [@kronmal_spurious_1993].  Unlike a ratio, maintaining properties of each lipid measure is possible with a CFA latent construct and perhaps this measure can better inform analyses with this construct as either an outcome or covariate.

Prior to implementation of this approach it is important to investigate if this latent construct represented in the CFA has a good model fit. Poor model fit would indicate that the fitted latent construct does not match the observed data very well and does not serve as a proper latent variable within a numerical dyslipidemia construct. For example, if the indicators are not correlated with one another then this approach is not feasible. Nor is it feasible if the choice of indicators do not fit within a sound theoretical basis. Neither of these cases apply to lipids used in common lipid profiles, and these two requirements make a case for examining dyslipidemia as a latent factor in CFA with the intent of using it in the future either as an exposure or outcome.

Assuming the three measures provide a good source of model fit, a next step would be to evaluate if the form of the latent construct of dyslipidemia stays relatively constant across demographic groups. Similar form of the construct across different socio-demographic groups allows comparisons of construct means across these groups -- a basic outcome of interest. One common method used to assess this property is measurement invariance (MI) of the latent factor across demographic groups [@van_de_schoot_editorial:_2015-1]. Absent MI then comparing construct means across the demographic groups is more difficult to accomplish. An alternative approach in that resulting scenario would be to stratify analyses for make meaningful inference.

Although the CFA requires additional knowledge of structural equation modeling, the ability to obtain a finer ability to capture correlations between lipid and lipoprotein indicators and condense it into a single latent variable can provide a parsimonious alternative approach to characterize dyslipidemia. Our aim in this study is to assess the fit and test measurement invariance of a latent construct representing dyslipidemia in a U.S.-representative sample. In turn, this assessment and tests can provide evidence of the viability of this construct in future models incorporating dyslipidemia.

# Methods

# Results

### 2017-08-02 draft/ms-201705-draft.Rmd
initial commit for cfa draft

--- 2017-05-24 manuscript.Rmd
@@
 ---
-title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups: United States, 2003 to 2012"
+title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions"
 author: "Ann Von Holle, Kari E. North"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+date: '`r format(Sys.time(), "%d %B, %Y")`'
 output:
+  word_document:
+    toc: yes
   html_document:
-    toc: true # table of content true
-    depth: 3  # up to three depths of headings (specified by #, ## and ###)
-    number_sections: true  ## if you want number sections at each table header
-    theme: united  # many options for theme, this one is my favorite.
-    highlight: tango  # specifies the syntax highlighting style
-bibliography: lit.bib
-csl: ../../vancouver.csl
+    csl: ama.csl
+    depth: 3
+    highlight: tango
+    number_sections: yes
+    theme: united
+    toc: yes
+csl: ama.csl
+bibliography: bib1.bib
 ---
@@
 
+# Abstract
+
+Lipids and lipoproteins found in common profiles such as LDL-C, HDL-C and TG are used to assess dyslipidemia, a well-established risk factor for CVD.
+These measures are commonly evaluated across socio-demographic groups in a univariate format -- one at a time.
+Assessing dyslipidemia as a multivariate construct is possible, and perhaps more meaningful, in the context of SEM, but no thorough assessment exists in the literature.
+
+To assess the multivariate construct, we will use confirmatory factor analysis (CFA) to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG.
+In turn, we will assess measurement invariance (MI)  of the construct across demographic characteristics of race/ethnicity and sex in a United States representative sample of children ages 9 to 12 years.
+
+Proper model fit will guide viability of this dyslipidemia construct in analyses.
+If measurement invariance across racial/ethnic or gender groups exists then there would also be no evidence of differences in the construct itself across these groups and allows for comparisons in means of the construct.
+Also, similarities across groups also implies the measured lipid indicators have similar pair-wise correlations.
+Conversely, measurement variance is an indicator that this latent factor is not the same across groups and should be treated as such with stratification.
+
+Results:
+
+We were not able to find evidence of measurement invariance (MI).
+Tests for MI across racial/ethnic or gender groups are significant indicating that these groups have different multivariate shapes of the three lipid/lipoproteins and the means of these constructs cannot be compared across groups.
+Within strata of gender or race, the construct has good fit via fit statistics making it a valid construct.
+
+Conclusion:
+
+Analyses demonstrate this construct is a parsimonious way to evaluate dyslipidemia and can provide an alternate approach to treatment of dyslipidemia in structural equation models.
+However, evaluation of this construct across socio-demographic groups such as race/ethnicity and gender provides evidence of a different multivariate structure requiring stratification.
+
 # Introduction
 
-Dyslipidemia is often characterized through measures of triglycerides, high-density lipoprotein cholesterol (HDL-C), and low-density cholesterol (LDL-C) from a common lipid panel.
+Dyslipidemia is often characterized through measures of triglycerides (TG), high-density lipoprotein cholesterol (HDL-C), and low-density cholesterol (LDL-C) from a common lipid panel.
 These measures are then analyzed in research on a univariate basis or with ratios.
@@
 
-Assuming the three measures provide a good source of model fit, a next step would be to evaluate if the form of the latent construct of dyslipidemia stays relatively constant across demographic groups.
+Assuming the three measures provide a good source of model fit, a next step would be to evaluate if the form of the latent construct of dyslipidemia stays relatively constant across demographic groups including race/ethnicity, age and gender.
 Similar form of the construct across different socio-demographic groups allows comparisons of construct means across these groups -- a basic outcome of interest.
@@
 
+National Health and Nutrition Examination Survey (NHANES) is a nationwide representative complex survey sample in the United States that assesses population health.
+Lipid measures are among the data collected in the physical exam [@_nhanes_2017-3].
+These analyses include NHANES samples from 2003-2012, spanning ten years.
+TC and TG were obtained via from venous samples with enzymatic methods [@_nhanes_2017-1; @carroll_trends_2012].
+HDL were were measured with a direct immunoassay method [@_nhanes_2017], LDL-C was determined from the Friedewald equation.
+We included children ages 12 to 19 years in the sample and further subsetted the sample to include the largest racial/ethnic groups in the sample: non-Hispanic White, non-Hispanic Black, and Mexican American.
+Also, those reporting taking lipid lowering medication were excluded from the sample.
+After these exclusions there were xx individuals in the complex survey sample.
+
+The primary aim is to create a latent construct of dyslipidemia so we conducted a CFA by specifying one latent variable, dyslipidemia, with three indicators: LDL-C, HDL-C and TG.
+We specified three factor loadings, one for each indicator, and 3 error variances creating a ‘just identified’ model with zero degrees of freedom [@brown_confirmatory_2015].
+All variables required for complex survey sampling were used in the analyses, including the sampling weights, strata and clustering information.
+We used the Comparative Fit Index, (CFI), Bayesian Information Criterion (BIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
+BMI and age are associated with lipid levels and these two factors were allowed to covary with the latent lipid construct to improve model fit.
+
+Measurement invariance was determined by specifying 1) configural, 2) metric and 3) scalar models [@_mplus_2017; @van_de_schoot_checklist_2012].
+Socioeconomic groups included age (12-13, 14-15, 16-17, 18-19), gender and racial/ethnic groups.
+The configural model allows the factor loadings and intercepts to vary across the groups; the metric model holds the factor loading the same across groups; the scalar model holds both the intercept and factor loadings to be the same across groups.
+To test differences between each of the models we used the chi-square test with a scaling correction factor.
+If significant, then there was evidence that the more complex model included significant terms that should remain in the model [@brown_confirmatory_2015].
+If not significant then we kept the model with fewer parameters.
+
+One aspect that was not possible to incorporate into the model was the method to subset complex survey data given the sampling weights.
+We did sensitivity tests to determine any bias in ignoring that assumption made on the model parameter estimation and did not detect any difference in effect, considered negligible in substantive meaning.
+
+We used Mplus version [@muthen_mplus_1998] for the CFA analysis and R [@r_core_team_r:_2017] for other data handling tasks.
+
 # Results
+
+For each of the groupings, race/ethnicity and gender, we determined model fit across groups by conducting invariance tests with chi-square tests of three different models against each other.
+We determined the best fitting model by selecting the most parsimonious model that was no different, i.e.
+non-significant, from the more complex model.
+
+When comparing gender, racial/ethnic, and age groups, the metric model, allowing intercepts to vary but fixed factor loadings, was not significantly different (Chi-Square=5.6, 2 df, p<0.06; Chi-Square=11.6, 8 df, p>0.16; Chi-Square=8.8, 4 df, p>0.06, respectively) from the configural model, which allowed both loading and intercepts to vary, indicating similarity of factor loadings.
+On the other hand, the scalar model, fixing both intercepts and factor loadings was different from the configural model (Chi-Square=71.1, 5 df, p<0.0001; Chi-Square=189.2, 20 df, p<0.0001; Chi-Square=86.5, 10 df, p<0.0001, respectively), indicating a difference in factor intercepts.
+Evidence for different factor intercepts is further supported by a significant test of the scalar versus metric model (Chi-Square=68.8, 3 df, p<0.0001; Chi-Square=198.8, 12 df, p<0.0001, Chi-Square=82.0, 6 df, p<0.0001, respectively).
+Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts.
+[Due to these results it is worth noting that a strict interpretation of factor means across the groups is not possible given the different intercepts.
+Add this part to discussion below]
+
+Model fit information by groups is in table xx. We found xx.
+
+Parameters for the best fitting model are in table xx. We found xx.
+
+# Discussion
+
+# Conclusion

### 2017-08-21 mplus-results-manuscript.Rmd
initial commit for manuscript results

---
title: "Mplus Results"
author: "Ann Von Holle"
date: '`r format(Sys.time(), "%d %B, %Y")`'
bibliography: lit.bib
output:
  html_document:
        toc: yes
        toc_float: yes
        number_sections: true
---

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}

library(knitr)
opts_chunk$set(fig.width=10, fig.height=10, 
               fig.keep = 'all',
               fig.align="left", 
               echo=F, results='hide',
               warning=FALSE, message=FALSE, comment=NA)

library("MplusAutomation")
library(reshape2)

library(data.table)
library(plyr)
library(kableExtra)

library(ztable)
options(ztable.type="html")
#options(ztable.type="viewer") # for debugging ztable

```

```{r}

# Read in values
# -------------------

# set directiory to spot where Mplus output for configure testing is stored
setwd("~/../Dropbox/unc.grad.school/my-papers/ms-201608-2/ms-cfa-lipids/mplus/mplus-configural testing")

# Read in information from the .out files
# read in lgmm data info 
output = readModels()
names(output[[1]])
names(output[[1]]$parameters) # get names of different param output.

# get fit summaries from output files
df.sums = lapply(output, function(x) x$summaries)
df.sums[[1]] # check

# combine all model summaries across different groupings into one data frame
combine.sums = do.call("rbind.fill", df.sums)
head(combine.sums)

# make a unique id
combine.sums$id.model = rep(paste0(names(output)),
                              sapply(output, function(x) 
                                nrow(data.frame(x$summaries))))
table(combine.sums$id.model) # check

# Make identifying information for the parameters
handle1 = function(dat) {
dat = within(dat, {
      # set up labels for outcome and sex of child groups (for facetting below)
   groups = ifelse(grepl("groups.age", id.model)==T, "Age",
                 ifelse(grepl(".groups.re.", id.model)==T, "Race/Ethnicity", 
                        ifelse(grepl(".groups.sex.", id.model)==T, "Gender",
                               ifelse(grepl(".groups.year.", id.model)==T, "Time", NA))))
    })
}

combine.sums = handle1(combine.sums)
table(combine.sums$groups) # check

# --------------------------------------------------
# --------------------------------------------------

# get parameters from output files -- for metric models 
# -------------------------------------------------------
# TO DO: need to get standardized estimates
df.params = lapply(output, function(x) data.frame(x$parameters$unstandardized$METRIC.MODEL))
nrow(df.params[[2]])

# Combine all param info across different groupsings into one data frame
combine.params = do.call("rbind.fill", df.params)
nrow(combine.params)

# make a unique id
combine.params$id.model = rep(paste0(names(output)),
                              sapply(output, function(x) 
                                nrow(data.frame(x$parameters$unstandardized$METRIC.MODEL))))
table(combine.params$id.model)

combine.params = handle1(combine.params)
table(combine.params$groups) # check

```

```{r}

# make a table of fit statistics
names(combine.sums)
class(combine.sums)

fits = c("ChiSqM_Value", "ChiSqM_DF",
         "LL", "CFI", "aBIC", "RMSEA_Estimate")

sums.dt = data.table(combine.sums)
wide.fit <- data.table::dcast(sums.dt[sums.dt$Model=="METRIC MODEL", with=T],
                  groups ~ Model , 
                  value.var = fits)

wide.fit

```

```{r, results='markup'}

#dim(wide.fit)
wide.fit2 = wide.fit
colnames(wide.fit2) = c(rep(c(""),7))

kable(wide.fit2, 
      booktabs=T,
      format="html",
      caption='Goodness of fit') %>%
    kable_styling(bootstrap_options = c("striped")) %>%
  add_header_above(c("Groups"=1,
                     "Chi-Square"=1,
                     "DF"=1,
                     "LL"=1,
                     "CFI"=1,
                     "aBIC"=1,
                     "RMSEA"=1)) #%>%
  # group_rows("Weight", 1, 7) %>%
  # group_rows("Height", 8, 14) %>%
  # group_rows("WFL", 15, 21)

```

### 2017-08-23 mplus-results-manuscript.Rmd
update to cfa tables

@@
 names(output[[1]]$parameters) # get names of different param output.
+```
 
+```{r}
 # get fit summaries from output files
@@
 nrow(df.params[[2]])
+df.params[[1]]
 
@@
 
+# Need to figure out a way to add the groups for each of there unique combos of 
+# paramHeader and param.
+# order of groups: age (3), re (5), sex (2) and year (5)
+
+# note that the f1.by values are duplicated are the same for factor loadings
+# because they are held constant across groups.
+# for age: 1 = 12-13 2 = 14-16 3 = 17-19
+# for race/ethnicity:  1 = MA 2 = OH 3 = NHW 4 = NHB 5 = OTH
+# for sex of child: 1 = male 0 = female
+# for year: 2003 = y2003 2005 = y2005 2007 = y2007 2009 = y2009 2011 = y2011
+
+names(output)
+
+names(combine.params)
+table(combine.params$unstandardized.Group)
+nrow(combine.params)
+head(combine.params)
+
+length(unique(paste0(combine.params$unstandardized.paramHeader, 
+                     combine.params$unstandardized.param)))
+
+combine.params$group = rep(paste0(names(output)),
+                              sapply(output, function(x) 
+                                nrow(data.frame(x$parameters$unstandardized$METRIC.MODEL))))
+
 combine.params = handle1(combine.params)
 table(combine.params$groups) # check
+
+nrow(combine.params[combine.params$groups=="Age",])
 
@@
 sums.dt = data.table(combine.sums)
-wide.fit <- data.table::dcast(sums.dt[sums.dt$Model=="METRIC MODEL", with=T],
-                  groups ~ Model , 
+wide.fit <- data.table::dcast(sums.dt,
+                  groups + Model ~ . , 
                   value.var = fits)
@@
 wide.fit2 = wide.fit
-colnames(wide.fit2) = c(rep(c(""),7))
+wide.fit2$Model = factor(wide.fit2$Model, 
+                         labels=c("Configural",
+                                  "Metric",
+                                  "Scalar"))
+kable(wide.fit2[,-1], 
+      booktabs=T,
+      format="html",
+      caption='Goodness of fit',
+      col.names=c("Model", "ChiSq", "df", "LL", "CFI", "aBIC", "RMSEA")) %>%
+    kable_styling(bootstrap_options = c("striped")) %>%
+  group_rows("Age", 1, 3) %>%
+  group_rows("Gender", 4, 6) %>%
+  group_rows("Race/ethnicity", 7, 9) %>%
+  group_rows("Time", 10, 12)
 
-kable(wide.fit2, 
+```
+
+```{r}
+
+# make a table of parameters
+names(combine.params)
+head(combine.params$groups)
+
+sums.dt = data.table(combine.sums)
+wide.params <- data.table::dcast(setDT(combine.params),
+                  groups +  unstandardized.Group + unstandardized.paramHeader + unstandardized.param  ~ .,
+                  value.var = c("unstandardized.est"))
+
+# note that the f1.by values are duplicated are the same for factor loadings
+# because they are held constant across groups.
+# for age: 1 = 12-13 2 = 14-16 3 = 17-19
+# for race/ethnicity:  1 = MA 2 = OH 3 = NHW 4 = NHB 5 = OTH
+# for sex of child: 1 = male 0=female
+# for year: 003 = y2003 2005 = y2005 2007 = y2007 2009 = y2009 2011 = y2011
+
+wide.params
+
+num.rows.p = table(wide.params$groups)
+num.rows.p
+num.rows.p.c = cumsum(num.rows.p)
+
+sub.wide.p = wide.params[wide.params$unstandardized.paramHeader=="Intercepts", ]
+num.rows.p.c = cumsum(table(sub.wide.p$groups))
+
+```
+
+```{r, results='markup'}
+
+kable(wide.params[wide.params$unstandardized.paramHeader=="Intercepts", -1], 
       booktabs=T,
@@
     kable_styling(bootstrap_options = c("striped")) %>%
-  add_header_above(c("Groups"=1,
-                     "Chi-Square"=1,
-                     "DF"=1,
-                     "LL"=1,
-                     "CFI"=1,
-                     "aBIC"=1,
-                     "RMSEA"=1)) #%>%
-  # group_rows("Weight", 1, 7) %>%
-  # group_rows("Height", 8, 14) %>%
-  # group_rows("WFL", 15, 21)
+  group_rows("Age", 1, num.rows.p[1]) %>%
+  group_rows("Gender", num.rows.p.c[1]+1, num.rows.p.c[2]) %>%
+  group_rows("Race/Ethnicity", num.rows.p.c[2]+1, num.rows.p.c[3]) %>%
+  group_rows("Time", num.rows.p.c[3]+1, num.rows.p.c[4])
 

### 2017-08-23 draft/ms-201705-draft.Rmd
updates to cfa ms

@@
 ---
-title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions"
+title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups -- United States, 2003 to 2012
 author: "Ann Von Holle, Kari E. North"
@@
 output: 
-  word_document:
-    toc: yes
   html_document:
-    csl: ama.csl
-    depth: 3
-    highlight: tango
-    number_sections: yes
-    theme: united
-    toc: yes
+    toc: true # table of content true
+    depth: 3  # up to three depths of headings (specified by #, ## and ###)
+    number_sections: true  ## if you want number sections at each table header
+    theme: united  # many options for theme, this one is my favorite.
+    highlight: tango  # specifies the syntax highlighting style
     csl: ama.csl
 bibliography: bib1.bib
+csl: ama.csl
 ---
@@
 All variables required for complex survey sampling were used in the analyses, including the sampling weights, strata and clustering information.
-We used the Comparative Fit Index, (CFI), Bayesian Information Criterion (BIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
+We used the Comparative Fit Index, (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
 BMI and age are associated with lipid levels and these two factors were allowed to covary with the latent lipid construct to improve model fit.
@@
 Evidence for different factor intercepts is further supported by a significant test of the scalar versus metric model (Chi-Square=68.8, 3 df, p<0.0001; Chi-Square=198.8, 12 df, p<0.0001, Chi-Square=82.0, 6 df, p<0.0001, respectively).
-Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts.
+Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts -- the metric model.
 [Due to these results it is worth noting that a strict interpretation of factor means across the groups is not possible given the different intercepts.
@@
 
-Model fit information by groups is in table xx. We found xx.
+Model fit information (Table 1) was satisfactory for the analyses by gender, racial/ethnic and age groups.
+First, RMSEA was below 0.04 in all three cases.
+Second, CFI was above 0.98.
+Third, aBIC was lowest for the metric model for all three models with the exception of the gender groups, and in this scenario, the difference between the configural and metric model was less than 4 points, considered to be a negligible difference.
+All of these measures of model fit point towards the metric model having satisfactory model fit and useful for further exploration of the model parameters.
 
-Parameters for the best fitting model are in table xx. We found xx.
+Parameters of interest for the best fitting model are the intercepts since they are the ones that are allowed to vary across groups in the metric specification of the model.
+In table 2, the intercepts for the confirmatory factor representing dyslipidemia are described by groups within each separate model.
+Within the model that separates by age, the three older age groups appear to have shift in the center of the factor in the negative direction for HDL, with standardized intercept resting at -0.163 for the 14 to 16 and 17 to 19 year age groups.
+For the middle age group, ages 14 to 16, LDL and TG are shifted further to the negative direction and the 17 to 19 year old group are shifted more in the positive direction when compared to the youngest age group 12 to 13 years of age.
+
+When examining the model with gender groups, the males appear to have a shift in the negative direction for HDL and towards a higher value of TG compared to females.
+Finally, the racial/ethnic groups also have differences in their intercepts.
+Using the Mexican American (MA) group as a referent, the non-Hispanic Blacks (NHB) have a center for the three-dimensional factor that is shifted more towards a higher HDL and LDL along with a lower TG.
+The non-Hispanic White group does not appear to have as large a difference as the NHB group.
+The other Hispanic (OH) group appears to have a larger negative shift in LDL and TG as well as the “Other” racial/ethnic group displaying characteristics similar to the NHB but not as extreme.
+
+Finally, factor loadings are not allowed to vary across groups in the metric model.
+Factor loadings indicate the degree to which the measured correlation between the indicators are correlated with the dyslipidemia factor.
+TG has the strongest correlation with the factor, at a correlation exceeding 0.71.
+HDL has the second strongest association -- this time a negative one -- around -0.52.
+Finally, LDL is the least correlated with the factor with a correlation at 0.39 for this sample.
+Any standardized factor loading below 0.3 indicates need to remove that indicator, and none of the lipid indicators fit that description.
+These values culminate in a factor indicating that a one unit increase in the factor translates to adverse levels of lipids with a decline in HDL and increases in TG and LDL.
 
@@
 # Conclusion
+
+# Tables
+
+```{r chi, child='../mplus-results-manuscript.Rmd'}
+
+```

### 2017-08-23 mplus-results-manuscript.Rmd
updates to cfa ms

@@
 
-# set directiory to spot where Mplus output for configure testing is stored
+# set directiory to spot where Mplus output for configural testing is stored
 setwd("~/../Dropbox/unc.grad.school/my-papers/ms-201608-2/ms-cfa-lipids/mplus/mplus-configural testing")
@@
 
-# Make identifying information for the parameters
+# Make identifying information for the summaries
 handle1 = function(dat) {
@@
 table(combine.sums$groups) # check
-
-# --------------------------------------------------
-# --------------------------------------------------
-
-# get parameters from output files -- for metric models 
-# -------------------------------------------------------
-# TO DO: need to get standardized estimates
-df.params = lapply(output, function(x) data.frame(x$parameters$unstandardized$METRIC.MODEL))
-nrow(df.params[[2]])
-df.params[[1]]
-
-# Combine all param info across different groupsings into one data frame
-combine.params = do.call("rbind.fill", df.params)
-nrow(combine.params)
-
-# make a unique id
-combine.params$id.model = rep(paste0(names(output)),
-                              sapply(output, function(x) 
-                                nrow(data.frame(x$parameters$unstandardized$METRIC.MODEL))))
-table(combine.params$id.model)
-
-# Need to figure out a way to add the groups for each of there unique combos of 
-# paramHeader and param.
-# order of groups: age (3), re (5), sex (2) and year (5)
-
-# note that the f1.by values are duplicated are the same for factor loadings
-# because they are held constant across groups.
-# for age: 1 = 12-13 2 = 14-16 3 = 17-19
-# for race/ethnicity:  1 = MA 2 = OH 3 = NHW 4 = NHB 5 = OTH
-# for sex of child: 1 = male 0 = female
-# for year: 2003 = y2003 2005 = y2005 2007 = y2007 2009 = y2009 2011 = y2011
-
-names(output)
-
-names(combine.params)
-table(combine.params$unstandardized.Group)
-nrow(combine.params)
-head(combine.params)
-
-length(unique(paste0(combine.params$unstandardized.paramHeader, 
-                     combine.params$unstandardized.param)))
-
-combine.params$group = rep(paste0(names(output)),
-                              sapply(output, function(x) 
-                                nrow(data.frame(x$parameters$unstandardized$METRIC.MODEL))))
-
-combine.params = handle1(combine.params)
-table(combine.params$groups) # check
-
-nrow(combine.params[combine.params$groups=="Age",])
-
 ```
@@
 
-```{r, results='markup'}
+## Model fit statistics
+
+```{r fit, results='markup'}
 
@@
 
+# Read in parameter estimates
+# ----------------------------
+
+# set directiory to spot where Mplus output for metric standardized param estimates are
+setwd("~/../Dropbox/unc.grad.school/my-papers/ms-201608-2/ms-cfa-lipids/mplus/mplus-configural testing/no-test")
+
+# Read in information from the .out files
+# read in lgmm data info 
+output.p = readModels()
+names(output.p[[1]])
+names(output.p[[1]]$parameters) # get names of different param output.
+```
+
+```{r}
+
+# --------------------------------------------------
+# --------------------------------------------------
+
+# get parameters from output files -- for metric models 
+# -------------------------------------------------------
+
+df.params = lapply(output.p, function(x) {
+  data.frame(x$parameters$stdyx.standardized)
+} )
+
+nrow(df.params[[2]])
+df.params[[1]]
+
+# Combine all param info across different groupsings into one data frame
+combine.params = do.call("rbind.fill", df.params)
+nrow(combine.params)
+names(output.p)
+
+# make a unique id
+combine.params$id.model = rep(paste0(names(output.p)),
+                              sapply(output.p, function(x) 
+                                nrow(data.frame(x$parameters))))
+table(combine.params$id.model)
+
+# Need to figure out a way to add the groups for each of there unique combos of 
+# paramHeader and param.
+# order of groups: age (3), re (5), sex (2) and year (5)
+
+# note that the f1.by values are duplicated are the same for factor loadings
+# because they are held constant across groups.
+# for age: 1 = 12-13 2 = 14-16 3 = 17-19
+# for race/ethnicity:  1 = MA 2 = OH 3 = NHW 4 = NHB 5 = OTH
+# for sex of child: 1 = male 0 = female
+# for year: 2003 = y2003 2005 = y2005 2007 = y2007 2009 = y2009 2011 = y2011
+
+combine.params = handle1(combine.params)
+table(combine.params$groups) # check
+
+```
+
+```{r}
+
 # make a table of parameters
+head(combine.params)
+
 names(combine.params)
@@
 
-sums.dt = data.table(combine.sums)
-wide.params <- data.table::dcast(setDT(combine.params),
-                  groups +  unstandardized.Group + unstandardized.paramHeader + unstandardized.param  ~ .,
-                  value.var = c("unstandardized.est"))
+wide.params <- data.table::dcast(setDT(combine.params[combine.params$paramHeader=="Intercepts",]),
+                  groups + Group  ~ param,
+                  value.var = c("est"))
 
-# note that the f1.by values are duplicated are the same for factor loadings
-# because they are held constant across groups.
+# standardized factor loadings
+wide.params.f <- data.table::dcast(setDT(combine.params[combine.params$paramHeader=="F1.BY",]),
+                  groups + Group  ~ param,
+                  value.var = c("est"))
+
+# groups:
 # for age: 1 = 12-13 2 = 14-16 3 = 17-19
@@
 
-sub.wide.p = wide.params[wide.params$unstandardized.paramHeader=="Intercepts", ]
-num.rows.p.c = cumsum(table(sub.wide.p$groups))
-
 ```
 
-```{r, results='markup'}
+## Standardized Intercepts
 
-kable(wide.params[wide.params$unstandardized.paramHeader=="Intercepts", -1], 
+```{r int, results='markup'}
+
+kable(wide.params[, -1], 
       booktabs=T,
       format="html",
-      caption='Goodness of fit') %>%
+      caption='Standardized Intercepts') %>%
     kable_styling(bootstrap_options = c("striped")) %>%
@@
 ```
+
+# Standardized Factor Loadings
+
+```{r factorload}
+
+kable(wide.params.f[, -1], 
+      booktabs=T,
+      format="html",
+      caption='Standardized Intercepts') %>%
+    kable_styling(bootstrap_options = c("striped")) %>%
+  group_rows("Age", 1, num.rows.p[1]) %>%
+  group_rows("Gender", num.rows.p.c[1]+1, num.rows.p.c[2]) %>%
+  group_rows("Race/Ethnicity", num.rows.p.c[2]+1, num.rows.p.c[3]) %>%
+  group_rows("Time", num.rows.p.c[3]+1, num.rows.p.c[4])
+
+```

### 2017-09-07 draft/ms-201705-draft.Rmd
update to draft

@@
 ---
-title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups -- United States, 2003 to 2012
+title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across
+  time and race-ethnic groups -- United States, 2003 to 2012
 author: "Ann Von Holle, Kari E. North"
 date: '`r format(Sys.time(), "%d %B, %Y")`'
+output:
+  html_document:
+    csl: ama.csl
+    depth: 3
+    highlight: tango
+    number_sections: yes
+    theme: united
+    toc: yes
+  word_document:
+    toc: yes
+csl: ama.csl
+bibliography: bib1.bib
+---
+---
+title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups: United States, 2003-2012
+author: "Ann Von Holle, Kari E. North"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
@@
 Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts -- the metric model.
-[Due to these results it is worth noting that a strict interpretation of factor means across the groups is not possible given the different intercepts.
-Add this part to discussion below]
 
 Model fit information (Table 1) was satisfactory for the analyses by gender, racial/ethnic and age groups.
-First, RMSEA was below 0.04 in all three cases.
+First, RMSEA was below 0.05 in all three cases.
 Second, CFI was above 0.98.
 Third, aBIC was lowest for the metric model for all three models with the exception of the gender groups, and in this scenario, the difference between the configural and metric model was less than 4 points, considered to be a negligible difference.
-All of these measures of model fit point towards the metric model having satisfactory model fit and useful for further exploration of the model parameters.
+All of these measures of model fit point towards the metric model having satisfactory model fit and useful for further exploration of the model parameters [@brown_confirmatory_2015].
 
@@
 Finally, factor loadings are not allowed to vary across groups in the metric model.
-Factor loadings indicate the degree to which the measured correlation between the indicators are correlated with the dyslipidemia factor.
-TG has the strongest correlation with the factor, at a correlation exceeding 0.71.
+Factor loadings indicate the degree to which the measured correlation between the indicators are correlated with the dyslipidemia factor concurrently separating out the measurement error -- an advantage of CFA.
+Examining these association shows that TG has the strongest correlation with the factor, at a correlation exceeding 0.71.
 HDL has the second strongest association -- this time a negative one -- around -0.52.
@@
 Any standardized factor loading below 0.3 indicates need to remove that indicator, and none of the lipid indicators fit that description.
-These values culminate in a factor indicating that a one unit increase in the factor translates to adverse levels of lipids with a decline in HDL and increases in TG and LDL.
+These values culminate in an interpretation of a one unit increase in the factor equating with an increase in adverse levels of lipids: declining HDL and increasing TG and LDL.
 
@@
 
-# Conclusion
+The purpose of this paper was to examine the use of CFA as a measure of dyslipidemia.
+First, we evaluated measures of model fit of a CFA with three common lipid measures serving as observed indicators in the model: HDL-C, LDL-C and TG.
+Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity and age -- a common area of comparisons for lipid measures.
+Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept.
+MI did not hold for different racial/ethnic nor to age groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.
 
-# Tables
+## Measurement Invariance
 
-```{r chi, child='../mplus-results-manuscript.Rmd'}
+In evaluating the three different types of models across subgroups for each of the following domains: age, race/ethnicity and sex of child, the metric model was the best of the three for all domains.
+The configural model, in which all factor loadings and means of the dyslipidemia factor were allowed to vary across groups, was not statistically significantly different from the metric model.
+Given this result, the metric model offers a parsimonious fit by fixing the factor loadings to be the same across groups.
+Furthermore, the metric model was different from the scalar model, adding more evidence to support the choice of the metric model.
 
-```
+Similar factor loadings across the subgroups implies correlations between the three lipid indicators are the same across the subgroups within each domain.
+This finding may have relevance for lipid ratio measures such as HDL to TG, a commonly used metric.
+For example, a higher factor score for a construct with positive HDL-C factor loadings combined with negative LDL-C and TG factor loadings would mean on average a simultaneous increase of HDL and decrease in LDL-C and TG, equivalent to a higher HDL-C to LDL-C ratio.
+This interpretation implies higher ratios are analogous to higher factor score as well as lower ratios having lower factor scores.
+
+Although the factor loadings for the construct are similar, this finding does not imply metric invariance holds for these subgroups.
+In fact, the measurement invariance testing does not support comparing factor means across subgroups within each domain.
+Metric invariance implies the centers of this multidimensional construct are different across the subgroups.
+For example, when the factor is equal to zero for all subgroups, the estimated indicator (HDL-C, LDL-C and TG) are all different depending on the subgroup you choose.
+This result has implications regarding use of mean factor scores across subgroups.
+Some researchers tends to obtain factor scores and use them as observed entities, then comparing them across subgroups.
+These analyses demonstrate the perils in this approach with scientifically meaningless results given the evidence of scale differences.
+The existence of measurement non-invariance supports different dyslipidemia constructs for each of the selected subgroups for race/ethnicity, age and gender.
+As such, comparisons in factor score means have different meanings and are not substantively meaningful.
+Although this finding remains restricted to this example, it pertains to a nationwide representative sample of children in the United States and provides an important reminder when conducting analyses in other samples.
+
+## Fit
+
+Finally, having chosen the metric model for all of the three domains, we determined that model fit diagnostics were within normal bounds of adequate to good fit and as such motivates further use of this construct.
+Use of this construct would be an interesting supplement to univariate outcomes with a different and similarly meaningful interpretation.
+
+Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research.
+Lipids do not function independently of one another and a measure addressing their covariance is an asset.
+CFA leverages this characteristic of these measures to form a measure with different meaning from univariate indicators.
+Combining the observed correlations between the three common lipids alongside their common presentation in the literature and theoretical basis for dyslipidemia further justifies the use of CFA.
+
+This alternate construct can be an adjunct measure to univariate measures of lipids, which provides a means to assess the degree to which one measure increases as another measure either increase or decreases on a population level.
+Aside from the important caveat of measurement non-invariance, this work demonstrates the degree to which HDL-C, LDL-C and TG are correlated with one another on a population level.
+In turn this correlation can provide meaningful associations with other factors in downstream analyses.
+
+# References

### 2017-09-13 draft/ms-201705-draft.Rmd
update to manuscript and formatting changes.

--- 2017-08-02 draft/ms-201705-draft.Rmd
@@
 ---
-title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions"
+title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across
+  time and race-ethnic groups -- United States, 2003 to 2012
 author: "Ann Von Holle, Kari E. North"
@@
 output:
-  word_document:
-    toc: yes
   html_document:
-    csl: ama.csl
+    css: style.css
     depth: 3
@@
     number_sections: yes
-    theme: united
+    theme: readable
+    toc: yes
+  pdf_document:
+    toc: yes
+  word_document:
     toc: yes
@@
 
-Lipids and lipoproteins found in common profiles such as LDL-C, HDL-C and TG are used to assess dyslipidemia, a well-established risk factor for CVD.
+Dyslipidemia, a well-established risk factor for CVD, incorporates lipids and lipoproteins found in common profiles such as LDL-C, HDL-C and TG.
 These measures are commonly evaluated across socio-demographic groups in a univariate format -- one at a time.
-Assessing dyslipidemia as a multivariate construct is possible, and perhaps more meaningful, in the context of SEM, but no thorough assessment exists in the literature.
+Assessing dyslipidemia as a multivariate construct is possible, and perhaps also meaningful, in the context of confirmatory factor analysis (CFA), but no thorough assessment exists in the literature.
 
-To assess the multivariate construct, we will use confirmatory factor analysis (CFA) to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG.
+To assess the multivariate construct, we will use CFA to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG.
 In turn, we will assess measurement invariance (MI)  of the construct across demographic characteristics of race/ethnicity and sex in a United States representative sample of children ages 9 to 12 years.
 
-Proper model fit will guide viability of this dyslipidemia construct in analyses.
+Proper model fit will guide suitability of these indicators for dyslipidemia construct in analyses.
 If measurement invariance across racial/ethnic or gender groups exists then there would also be no evidence of differences in the construct itself across these groups and allows for comparisons in means of the construct.
@@
 
-The primary aim is to create a latent construct of dyslipidemia so we conducted a CFA by specifying one latent variable, dyslipidemia, with three indicators: LDL-C, HDL-C and TG.
+The primary aim is to create a latent construct of dyslipidemia.
+We conducted a CFA by specifying one latent variable, dyslipidemia, with three indicators: LDL-C, HDL-C and TG.
 We specified three factor loadings, one for each indicator, and 3 error variances creating a ‘just identified’ model with zero degrees of freedom [@brown_confirmatory_2015].
 All variables required for complex survey sampling were used in the analyses, including the sampling weights, strata and clustering information.
-We used the Comparative Fit Index, (CFI), Bayesian Information Criterion (BIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
+We used the Comparative Fit Index, (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
 BMI and age are associated with lipid levels and these two factors were allowed to covary with the latent lipid construct to improve model fit.
@@
 
-We used Mplus version [@muthen_mplus_1998] for the CFA analysis and R [@r_core_team_r:_2017] for other data handling tasks.
+We used Mplus version 8 [@muthen_mplus_1998] for the CFA analysis and R [@r_core_team_r:_2017; @hallquist_mplusautomation:_2017] for other data handling tasks.
 
@@
 
-For each of the groupings, race/ethnicity and gender, we determined model fit across groups by conducting invariance tests with chi-square tests of three different models against each other.
+For each of the groupings, race/ethnicity and gender, we determined model fit across groups by conducting Chi-Square invariance tests of three different models against each other.
 We determined the best fitting model by selecting the most parsimonious model that was no different, i.e.
@@
 Evidence for different factor intercepts is further supported by a significant test of the scalar versus metric model (Chi-Square=68.8, 3 df, p<0.0001; Chi-Square=198.8, 12 df, p<0.0001, Chi-Square=82.0, 6 df, p<0.0001, respectively).
-Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts.
-[Due to these results it is worth noting that a strict interpretation of factor means across the groups is not possible given the different intercepts.
-Add this part to discussion below]
+Thus the best fitting model across different gender, racial/ethnic, and age groups is one with similar factor loadings across the groups, but different factor intercepts -- the metric model.
 
-Model fit information by groups is in table xx. We found xx.
+Model fit information (Table 1) was satisfactory for the analyses by gender, racial/ethnic and age groups.
+First, RMSEA was below 0.05 in all three cases.
+Second, CFI was above 0.98.
+Third, aBIC was lowest for the metric model for all three models with the exception of the gender groups, and in this scenario, the difference between the configural and metric model was less than 4 points, considered to be a negligible difference.
+All of these measures of model fit point towards the metric model having satisfactory model fit and useful for further exploration of the model parameters [@brown_confirmatory_2015].
 
-Parameters for the best fitting model are in table xx. We found xx.
+CFA intercepts are parameters of interest for the best fitting model because they vary across groups in the metric specification of the model.
+In table 2, the intercepts for the confirmatory factor representing dyslipidemia are described by groups within each separate model.
+Within the model that separates by age, the three older age groups appear to have lower intercepts for HDL, with a standardized intercept resting at -0.163 for the 14 to 16 and 17 to 19 year age groups.
+For the middle age group, ages 14 to 16, LDL and TG intercepts are shifted further to the negative direction.
+In contrast, the 17 to 19 year old group have LDL and TG intercepts that are shifted closer to the origin when compared to the youngest age group.
+
+For gender groups, the males appear to have a shift in the negative direction for HDL and towards the origin for TG when compared to females.
+Finally, the racial/ethnic groups also have differences in their CFA intercepts.
+Using the Mexican American (MA) group as a referent, the non-Hispanic Blacks (NHB) have a positive intercept for HDL unlike the majority of the other groups and the largest negative shift in the intercept for TG compared to the other groups.
+The non-Hispanic White group appears to have similar intercept values compared to the MA group.
+Finally, the other Hispanic (OH) group appears to have the largest  negative shift in LDL-C.
+One can think of the intercept for each subgroup to be the mean lipid outcome for the average dyslipidemia factor -- in this case when the dyslipidemia factor is zero since it is mean-centered in these analyses.
+These differences in intercepts for the groups have consequences for intergroup comparisons as will be discussed.
+
+Finally, factor loadings are not allowed to vary across groups in the metric model.
+Standardized factor loadings without cross-loadings -- as in the specified analyses -- indicate the correlation between the dyslipidemia factor and the indicator while omitting measurement error.
+The TG indicator has the strongest correlation with the factor, at a correlation exceeding 0.71.
+HDL has the second strongest association -- around -0.52.
+Finally, LDL is the least correlated with the factor with a correlation at 0.39 for this sample.
+Any standardized factor loading below 0.3 indicates need to remove that indicator, and none of the lipid indicators fit that description.
+These values culminate in an interpretation of a one unit increase in the factor equating with an increase in unfavorable levels of lipids: declining HDL and increasing TG and LDL.
 
@@
 
-# Conclusion
+The purpose of this paper was to examine the use of CFA as a measure of dyslipidemia.
+First, we evaluated measures of model fit of a CFA with three common lipid measures serving as observed indicators in the model: HDL-C, LDL-C and TG.
+Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity and age -- common ground for comparisons.
+Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept.
+MI did not hold for different racial/ethnic nor to age groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.
+
+## Measurement Invariance
+
+In evaluating the three different types of models across subgroups for each of the following domains: age, race/ethnicity and sex of child, the metric model was the best of the three for all domains.
+The configural model, in which all factor loadings and means of the dyslipidemia factor were allowed to vary across groups, was not statistically significantly different from the metric model.
+Given this result, the metric model offers a parsimonious fit by fixing the factor loadings to be the same across groups.
+Furthermore, the metric model was different from the scalar model, adding more evidence to support the choice of the metric model.
+
+Testing of measurement invariance may have intriguing implications.
+First, there is no evidence for differences in factor loadings across any subgroups of for race/ethnicity, gender and age.
+This finding supports the idea of a universal system in which the three lipid indicators interact no matter the choice of sociodemographic subgroup.
+Confounding may play less of a part in this explanation than that for the adult population as exposure to smoking and lifestyle choices has had less time to exact a cumulative toll.
+Also, a similarity of factor loadings across subgroups may have relevance for lipid ratio measures such as HDL to TG, a commonly used metric.
+For example, a dyslipidemia factor with positive HDL-C factor loadings combined with negative LDL-C and TG factor loadings translates to a simultaneous increase of HDL and decrease in LDL-C and TG.
+In turn this factor interpretation could be an analog to HDL-C to TG ratio changes, implying  generalizability of these ratios across sociodemographic subgroups.
+No research has been done comparing these measures and would be of interest.
+
+Although the factor loadings for the construct are similar, we did not find evidence that metric invariance holds for these subgroups.
+Metric non-invariance implies at least one group has a different response to the lipid latent construct [@van_de_schoot_checklist_2012].
+For example, when the factor is equal to zero for all subgroups, the estimated mean indicator (HDL-C, LDL-C and TG), the intercept, is different depending on the subgroup you choose.
+These multivariate findings, especially of the African American group having lower TG and higher HDL supports prior findings of univariate mean differences across racial/ethnic groups [@sumner_ethnic_2009; @kit_prevalence_2015].
+
+Importantly, metric non-invariance for this factor restricts the use of mean factor scores across subgroups.
+Some researchers tend to obtain factor scores and use them as observed entities, followed by subgroup comparisons.
+These analyses demonstrate the perils in this approach for subgroup comparison of the dyslipidemia construct that result in scientifically meaningless estimates.
+Measurement non-invariance for all three domains supports different dyslipidemia constructs for each of the selected subgroups for race/ethnicity, age and gender.
+As such, comparisons in factor score means have different meanings and are not substantively meaningful.
+Although this finding remains restricted to this sample, it is of a nationwide representative sample of children in the United States and provides a cautionary tale of interpretation when conducting analyses in other samples.
+
+## Fit
+
+Finally, having chosen the metric model for all of the three domains, we determined that model fit diagnostics were within normal bounds of adequate to good fit and as such motivates further use of this construct.
+Use of this construct would be an interesting adjunct to univariate outcomes with a different and similarly meaningful interpretation.
+
+Strengths of these analyses include the representation of the 12 to 19 year old population in the United States.
+Limitations include the possibility of a nonlinear association between these lipid indicators, which a CFA does not account for and may represent a mis-specified model.
+[need to add more to this section]
+
+Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research.
+Lipids do not function independently of one another and a measure addressing their covariance is an asset.
+CFA leverages this characteristic of these measures to form a measure with different meaning from univariate indicators.
+Combining the observed correlations between the three common lipids alongside their common presentation in the literature and theoretical basis for dyslipidemia further justifies the use of CFA.
+
+Dyslipidemia as a multivariate construct can be an adjunct measure to univariate measures of lipids, which provides a means to assess the degree to which one measure increases as another measure either increase or decreases on a population level.
+Aside from the important caveat of measurement non-invariance, this work demonstrates the degree to which HDL-C, LDL-C and TG are correlated with one another on a population level.
+In turn this correlation can provide meaningful interpretation with other factors in downstream analyses.
+
+# References

### 2017-09-13 draft/ms-201705-draft.Rmd
updates

@@
 ---
-title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across
-  time and race-ethnic groups -- United States, 2003 to 2012
+title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups\\: United States 2003-2012"
 author: "Ann Von Holle, Kari E. North"
-date: '`r format(Sys.time(), "%d %B, %Y")`'
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
+  word_document:
+    toc: yes
   html_document:
+    csl: ama.csl
     css: style.css
@@
     theme: readable
-    toc: yes
-  pdf_document:
-    toc: yes
-  word_document:
     toc: yes
@@
 To assess the multivariate construct, we will use CFA to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG.
-In turn, we will assess measurement invariance (MI)  of the construct across demographic characteristics of race/ethnicity and sex in a United States representative sample of children ages 9 to 12 years.
+In turn, we will assess measurement invariance (MI)  of the construct across demographic characteristics of age, race/ethnicity and sex in a United States representative sample of children ages 12 to 19 years.
 
@@
 We used the Comparative Fit Index, (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
-BMI and age are associated with lipid levels and these two factors were allowed to covary with the latent lipid construct to improve model fit.
 
@@
 Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept.
-MI did not hold for different racial/ethnic nor to age groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.
+MI did not hold for different racial/ethnic, age or gender groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.
 
@@
 First, there is no evidence for differences in factor loadings across any subgroups of for race/ethnicity, gender and age.
-This finding supports the idea of a universal system in which the three lipid indicators interact no matter the choice of sociodemographic subgroup.
+This finding supports the idea of a universal system in which the three lipid indicators covary similarly no matter the choice of sociodemographic subgroup.
 Confounding may play less of a part in this explanation than that for the adult population as exposure to smoking and lifestyle choices has had less time to exact a cumulative toll.
@@
 Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research.
-Lipids do not function independently of one another and a measure addressing their covariance is an asset.
+In a biologic sense, lipids do not function independently of one another and an analytic measure addressing their covariance is an asset.
 CFA leverages this characteristic of these measures to form a measure with different meaning from univariate indicators.

### 2017-10-20 draft/ms-201705-draft.Rmd
updates

@@
 ---
-title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups\\: United States 2003-2012"
+title: 'Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups\: United States 2003\-2012'
 author: "Ann Von Holle, Kari E. North"
@@
 output:
-  word_document:
-    toc: yes
   html_document:
@@
     theme: readable
+    toc: yes
+  word_document:
     toc: yes
@@
 
-Dyslipidemia, a well-established risk factor for CVD, incorporates lipids and lipoproteins found in common profiles such as LDL-C, HDL-C and TG.
+Dyslipidemia, a well-established risk factor for cardiovascular disease (CVD), incorporates lipids and lipoproteins found in common profiles such as low density lipoprotein cholesterol (LDL-C), high density lipoprotein cholesterol (HDL-C) and triglycerides (TG).
 These measures are commonly evaluated across socio-demographic groups in a univariate format -- one at a time.
@@
 
-To assess the multivariate construct, we will use CFA to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG.
+To assess a multivariate construct of dyslipidemia, we will use CFA to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG as indicators.
 In turn, we will assess measurement invariance (MI) of the construct across demographic characteristics of age, race/ethnicity and sex in a United States representative sample of children ages 12 to 19 years.
 
-Proper model fit will guide suitability of these indicators for dyslipidemia construct in analyses.
-If measurement invariance across racial/ethnic or gender groups exists then there would also be no evidence of differences in the construct itself across these groups and allows for comparisons in means of the construct.
-Also, similarities across groups also implies the measured lipid indicators have similar pair-wise correlations.
+Proper model fit will guide suitability of these indicators for a dyslipidemia construct in analyses.
+If measurement invariance across racial/ethnic or gender groups exists then there would also be no evidence of differences in the construct itself across these groups and allows for comparisons in means of the construct [@van_de_schoot_editorial:_2015].
+Also, similarities across groups implies the measured lipid indicators have similar pair-wise correlations.
 Conversely, measurement variance is an indicator that this latent factor is not the same across groups and should be treated as such with stratification.
@@
 
-We were not able to find evidence of measurement invariance (MI).
+Measurement invariance (MI) did not hold in this sample.
 Tests for MI across racial/ethnic or gender groups are significant indicating that these groups have different multivariate shapes of the three lipid/lipoproteins and the means of these constructs cannot be compared across groups.
@@
 
-Analyses demonstrate this construct is a parsimonious way to evaluate dyslipidemia and can provide an alternate approach to treatment of dyslipidemia in structural equation models.
+Analyses demonstrate this construct is a alternate way to evaluate dyslipidemia using structural equation models.
 However, evaluation of this construct across socio-demographic groups such as race/ethnicity and gender provides evidence of a different multivariate structure requiring stratification.
@@
 Any changes in these values represented by a latent factor would be recognizable to lipid researchers and of value in research.
-What would most likely occur is for TG and LDL-C to be positively correlated and HDL-C negatively correlated with the prior two indicators.
+What would most likely occur in the construct is a positive correlation between TG and LDL-C and a negative correlation between HDL-C and the prior two indicators.
 A one unit change in the latent factor in this case could represent an increase in TG and LDL-C alongside a decrease in HDL-C given the well established correlations between these measures.
 
-A further motivation to examine dyslipidemia in a dimension reduction approach is to further develop the rationale behind the use of ratios when characterizing lipids.
-One could consider the correlations between the indicators to be an analog to ratios, with an additional ability to maintain variability for each measure.
+A further motivation to examine dyslipidemia in a dimension reduction approach is its similarity to ratios when characterizing lipids.
+One could consider the correlations between the indicators to be an analog to ratios, with an additional ability to retain the variability for each lipid indicator.
 For example, the use of HDL to TG ratios is increasing in the literature, but analyses using a ratio has limitations [@kuh_correlation_1955; @atchley_ratios_1978; @kronmal_spurious_1993] including failure to adequately adjust for each of the variables forming the ratio when used as a covariate in a linear regression [@kronmal_spurious_1993].
-Unlike a ratio, maintaining properties of each lipid measure is possible with a CFA latent construct and perhaps this measure can better inform analyses with this construct as either an outcome or covariate.
+Unlike a ratio, characterizing the variance of each lipid indicator is possible with a CFA latent construct.
 
@@
 For example, if the indicators are not correlated with one another then this approach is not feasible.
-Nor is it feasible if the choice of indicators do not fit within a sound theoretical basis.
+Nor is it feasible if the choice of indicators does not fit within a sound theoretical basis.
 Neither of these cases apply to lipids used in common lipid profiles, and these two requirements make a case for examining dyslipidemia as a latent factor in CFA with the intent of using it in the future either as an exposure or outcome.
@@
 One common method used to assess this property is measurement invariance (MI) of the latent factor across demographic groups [@van_de_schoot_editorial:_2015-1].
-Absent MI then comparing construct means across the demographic groups is more difficult to accomplish.
+Absent MI, comparing construct means across the demographic groups is more difficult to accomplish.
 An alternative approach in that resulting scenario would be to stratify analyses for make meaningful inference.
 
-Although the CFA requires additional knowledge of structural equation modeling, the ability to obtain a finer ability to capture correlations between lipid and lipoprotein indicators and condense it into a single latent variable can provide a parsimonious alternative approach to characterize dyslipidemia.
+Although the CFA requires additional knowledge of structural equation modeling, the ability to capture correlations between lipid and lipoprotein indicators, separately estimate measurement error, and condense the indicators into a single latent variable can provide a parsimonious alternative approach to characterize dyslipidemia.
 Our aim in this study is to assess the fit and test measurement invariance of a latent construct representing dyslipidemia in a U.S.-representative sample.
@@
 The primary aim is to create a latent construct of dyslipidemia.
-We conducted a CFA by specifying one latent variable, dyslipidemia, with three indicators: LDL-C, HDL-C and TG.
-We specified three factor loadings, one for each indicator, and 3 error variances creating a ‘just identified’ model with zero degrees of freedom [@brown_confirmatory_2015].
+We conducted a CFA by specifying one latent variable, named dyslipidemia, with three indicators: LDL-C, HDL-C and TG.
+We specified three factor loadings, one for each indicator, and three error variances creating a ‘just identified’ model with zero degrees of freedom [@brown_confirmatory_2015].
 All variables required for complex survey sampling were used in the analyses, including the sampling weights, strata and clustering information.
-We used the Comparative Fit Index, (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
+We used the Comparative Fit Index (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.
 
 Measurement invariance was determined by specifying 1) configural, 2) metric and 3) scalar models [@_mplus_2017; @van_de_schoot_checklist_2012].
-Socioeconomic groups included age (12-13, 14-15, 16-17, 18-19), gender and racial/ethnic groups.
+Socioeconomic groups included age (12-13, 14-15, 16-17, 18-19 years), gender and racial/ethnic groups (Mexican American, non-Hispanic Black, non-Hispanic White, other Hispanic, and ‘Other’).
 The configural model allows the factor loadings and intercepts to vary across the groups; the metric model holds the factor loading the same across groups; the scalar model holds both the intercept and factor loadings to be the same across groups.
@@
 
-For each of the groupings, race/ethnicity and gender, we determined model fit across groups by conducting Chi-Square invariance tests of three different models against each other.
-We determined the best fitting model by selecting the most parsimonious model that was no different, i.e.
+We determined model fit across groups by conducting Chi-Square invariance tests of three different models against each other in each of the three groupings: race/ethnicity, gender, and age.
+The best fitting model was the most parsimonious model that was no different, i.e.
 non-significant, from the more complex model.
@@
 CFA intercepts are parameters of interest for the best fitting model because they vary across groups in the metric specification of the model.
+One can think of the intercept for each subgroup to be the mean lipid outcome for the average dyslipidemia factor -- in this case when the dyslipidemia factor is zero since it is mean-centered in these analyses.
 In table 2, the intercepts for the confirmatory factor representing dyslipidemia are described by groups within each separate model.
+
 Within the model that separates by age, the three older age groups appear to have lower intercepts for HDL, with a standardized intercept resting at -0.163 for the 14 to 16 and 17 to 19 year age groups.
@@
 In contrast, the 17 to 19 year old group have LDL and TG intercepts that are shifted closer to the origin when compared to the youngest age group.
+Intercepts for gender groups demonstrate that males appear to have a shift of intercept in the negative direction for HDL and towards the origin for TG when compared to females.
 
-For gender groups, the males appear to have a shift in the negative direction for HDL and towards the origin for TG when compared to females.
 Finally, the racial/ethnic groups also have differences in their CFA intercepts.
@@
 The non-Hispanic White group appears to have similar intercept values compared to the MA group.
-Finally, the other Hispanic (OH) group appears to have the largest  negative shift in LDL-C.
-One can think of the intercept for each subgroup to be the mean lipid outcome for the average dyslipidemia factor -- in this case when the dyslipidemia factor is zero since it is mean-centered in these analyses.
+Lastly, the other Hispanic (OH) group appears to have the largest  negative shift in LDL-C.
 These differences in intercepts for the groups have consequences for intergroup comparisons as will be discussed.
 
-Finally, factor loadings are not allowed to vary across groups in the metric model.
+Factor loadings are not allowed to vary across groups in the metric model.
 Standardized factor loadings without cross-loadings -- as in the specified analyses -- indicate the correlation between the dyslipidemia factor and the indicator while omitting measurement error.
@@
 First, we evaluated measures of model fit of a CFA with three common lipid measures serving as observed indicators in the model: HDL-C, LDL-C and TG.
-Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity and age -- common ground for comparisons.
+Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity, age and gender -- common ground for comparisons.
 Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept.
-MI did not hold for different racial/ethnic, age or gender groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.
+MI did not hold for different racial/ethnic, age or gender groups implying that the factor means can not be meaningfully compared across these groups without special treatment [@van_de_schoot_facing_2013], which has implications for future use.
 
@@
 First, there is no evidence for differences in factor loadings across any subgroups of for race/ethnicity, gender and age.
-This finding supports the idea of a universal system in which the three lipid indicators covary similarly no matter the choice of sociodemographic subgroup.
+This finding supports the idea that the three lipid indicators covary similarly no matter the choice of sociodemographic subgroup.
 Confounding may play less of a part in this explanation than that for the adult population as exposure to smoking and lifestyle choices has had less time to exact a cumulative toll.
@@
 For example, when the factor is equal to zero for all subgroups, the estimated mean indicator (HDL-C, LDL-C and TG), the intercept, is different depending on the subgroup you choose.
-These multivariate findings, especially of the African American group having lower TG and higher HDL supports prior findings of univariate mean differences across racial/ethnic groups [@sumner_ethnic_2009; @kit_prevalence_2015].
+These multivariate findings, especially of the African American group having lower TG and higher HDL is in line with prior findings of univariate mean differences across racial/ethnic groups [@sumner_ethnic_2009; @kit_prevalence_2015].
 
 Importantly, metric non-invariance for this factor restricts the use of mean factor scores across subgroups.
-Some researchers tend to obtain factor scores and use them as observed entities, followed by subgroup comparisons.
-These analyses demonstrate the perils in this approach for subgroup comparison of the dyslipidemia construct that result in scientifically meaningless estimates.
+Using factor scores as observed entities followed by subgroup comparisons, given non-invariance, may result in scientifically meaningless estimates.
 Measurement non-invariance for all three domains supports different dyslipidemia constructs for each of the selected subgroups for race/ethnicity, age and gender.
-As such, comparisons in factor score means have different meanings and are not substantively meaningful.
 Although this finding remains restricted to this sample, it is of a nationwide representative sample of children in the United States and provides a cautionary tale of interpretation when conducting analyses in other samples.
@@
 Finally, having chosen the metric model for all of the three domains, we determined that model fit diagnostics were within normal bounds of adequate to good fit and as such motivates further use of this construct.
-Use of this construct would be an interesting adjunct to univariate outcomes with a different and similarly meaningful interpretation.
+Use of this construct would be an interesting adjunct to univariate outcomes with a different and meaningful interpretation.
 
 Strengths of these analyses include the representation of the 12 to 19 year old population in the United States.
-Limitations include the possibility of a nonlinear association between these lipid indicators, which a CFA does not account for and may represent a mis-specified model.
-[need to add more to this section]
-
-Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research.
+Also, measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research.
 In a biologic sense, lipids do not function independently of one another and an analytic measure addressing their covariance is an asset.
@@
 Combining the observed correlations between the three common lipids alongside their common presentation in the literature and theoretical basis for dyslipidemia further justifies the use of CFA.
+Limitations include the possibility of a nonlinear association between these lipid indicators, which a CFA does not account for and may represent a mis-specified model.
 
@@
 Aside from the important caveat of measurement non-invariance, this work demonstrates the degree to which HDL-C, LDL-C and TG are correlated with one another on a population level.
-In turn this correlation can provide meaningful interpretation with other factors in downstream analyses.
+In turn this correlation when captured via CFA can provide meaningful interpretation with other factors in downstream analyses.
 

### 2017-10-20 mplus-results-manuscript.Rmd
update to draft
updates

@@
 date: '`r format(Sys.time(), "%d %B, %Y")`'
-bibliography: lit.bib
 output:
   html_document:
+    number_sections: yes
     toc: yes
     toc_float: yes
-        number_sections: true
+  pdf_document:
+    toc: yes
+  word_document:
+    toc: yes
+bibliography: lit.bib
 ---
@@
 
+```{r factorload}
 # Standardized Factor Loadings
-
-```{r factorload}
 

### 2017-11-17 draft/ms-201705-draft.Rmd
update

@@
 ---
-title: 'Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups\: United States 2003\-2012'
+title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups: United States, 2003-2012"
 author: "Ann Von Holle, Kari E. North"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+date: '`r format(Sys.time(), "%d %B, %Y")`'
 output:
+  word_document:
+    toc: yes
   html_document:
@@
     theme: readable
-    toc: yes
-  word_document:
     toc: yes
@@
 
-Analyses demonstrate this construct is a alternate way to evaluate dyslipidemia using structural equation models.
+Analyses demonstrate this construct is an alternate way to evaluate dyslipidemia using structural equation models.
 However, evaluation of this construct across socio-demographic groups such as race/ethnicity and gender provides evidence of a different multivariate structure requiring stratification.
